---
title: "SCox_DS6372_Project2"
author: "Steven Cox"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 90
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Required Libraries
library(DataExplorer)
library(tidyverse)
library(dplyr)
library(lmtest)
library(car)
library(corrplot)
library(jtools)
library(sjPlot)
library(ResourceSelection)
library(ROCR)
library(pROC)
library(ggplot2)
library(naniar)
library(tidyverse)
library(readr)
library(dplyr)
library(ggplot2)
library(caret)

library(data.table)
library(funModeling)
library(caTools)


# set working directory
library(rstudioapi)
setwd(dirname(getSourceEditorContext()$path))
# rm(list = ls()))

library(rmarkdown)

# Convert Rmd to md
render("your_document.Rmd", output_format = "md_document")
```

## Section 1: Loading the Data

```{r load-data}
adult_raw <-
  read.csv("dataset/adult.data", header = FALSE, na.strings = "?")

colnames <-
  c(
    'age',
    'workclass',
    'similar_pop_count',
    'education_level',
    'years_education',
    'marital_status',
    'occupation',
    'household_role',
    'race',
    'gender',
    'capital_gain',
    'capital_loss',
    'hours_per_week',
    'native_country',
    'income'
  )

names(adult_raw) <- colnames

# Display the first few rows

glimpse(adult_raw)
```

### Remove Leading and Trailing Whitespaces

```{r remove whitespaces, warning=FALSE}
# Trim whitespace from all character columns
adult <- data.frame(lapply(adult_raw, function(x) {
  if (is.character(x)) trimws(x) else x
}))
```

### Refactor Income (Response Variable)

```{r refactored income, warning=FALSE}
# Define the new levels for income
adult$income <- adult$income %>%
  fct_recode(
    "<=50K" = "",
    "Less.50K" = "<=50K",
    "Greater.50K" = ">50K"
    )

summary(adult$income)
```

### Refactor Education Level
```{r refactor education_level, warning=FALSE}
# The new education level categories
adult$education_level <- adult$education_level %>%
  fct_recode(
    "highsch" = "10th",
    "highsch" = "11th",
    "highsch" = "12th",
    "primary" = "1st-4th",
    "primary" = "5th-6th",
    "secondary" = "7th-8th",
    "highsch" = "9th",
    "assoc" = "Assoc-acdm",
    "assoc" = "Assoc-voc",
    "undergrad" = "Bachelors",
    "phd" = "Doctorate",
    "highsch" = "HS-grad",
    "master" = "Masters",
    "no_edu" = "Preschool",
    "undergrad" = "Prof-school",
    "assoc" = "Some-college"
  )

adult$education_level <- adult$education_level %>%
  fct_relevel("no_edu", "primary", "secondary", "highsch", 
              "assoc", "undergrad", "master", "phd")

summary(adult$education_level)
```

### Refactor Working Class

```{r refactored workclass, warning=FALSE}

adult$workclass <- adult$workclass %>%
  fct_recode(
    "Unemployed" = "?",
    "FedGov" = "Federal-gov",
    "LocGov" = "Local-gov",
    "NeverWorked" = "Never-worked",
    "SelfEmpInc" = "Self-emp-inc",
    "SelfEmpNotInc" = "Self-emp-not-inc",
    "StateGov" = "State-gov",
    "Unemployed" = "Without-pay"
  )

summary(adult$workclass)
```

### Refactor Maritial Status

```{r refactored marital_status, warning=FALSE}
# The new marital status categories
adult$marital_status <- adult$marital_status %>%
  fct_recode(
    "divorced" = "Divorced",
    "married" = "Married-AF-spouse",
    "married" = "Married-civ-spouse",
    "married" = "Married-spouse-absent",
    "not_married" = "Never-married",
    "divorced" = "Separated",
    "widowed" = "Widowed"
  )
# recreate the levels
levels(adult$marital_status) <-
  c("divorce", "married", "notmarried", "widowed")

summary(adult$marital_status)
```

### Refactor Occupation

```{r refactored occupation, warning=FALSE}
# The new occupation levels
adult$occupation <- adult$occupation %>%
  fct_recode(
    "other" = "?",
    "clerical" = "Adm-clerical",
    "low_skill" = "Armed-Forces",
    "mid_skill" = "Craft-repair",
    "high_skill" = "Exec-managerial",
    "agriculture" = "Farming-fishing",
    "low_skill" = "Handlers-cleaners",
    "mid_skill" = "Machine-op-inspct",
    "low_skill" = "Other-service",
    "low_skill" = "Priv-house-serv",
    "high_skill" = "Prof-specialty",
    "high_skill" = "Protective-serv",
    "high_skill" = "Sales",
    "high_skill" = "Tech-support",
    "mid_skill" = "Transport-moving"
  )
levels(adult$occupation) <-
  c("other", "clerical", "lowskill", "midskill", "highskill", "agriculture")

summary(adult$occupation)
```

### Refactor Household Role

```{r refactored household_role, warning=FALSE}
# Define the new levels for household_role
adult$household_role <- adult$household_role %>%
  fct_recode(
    "husband" = "Husband",
    "tenant" = "Not-in-family",
    "relative" = "Other-relative",
    "depchild" = "Own-child",
    "single_occ" = "Unmarried",
    "wife" = "Wife"
  )

summary(adult$household_role)
```

### Refactor Native Country

```{r refactored native_region, warning=FALSE}
# Function to recode the values with trimmed spaces
recode_value <- function(old_level) {
  old_level <- trimws(old_level) # Trim leading and trailing whitespace
  for (category in names(new_levels)) {
    if (old_level %in% new_levels[[category]]) {
      return(category)
    }
  }
  return(NA) # Return NA if the education level does not match any category
}

# Define the new levels for native_region
new_levels <- list(
  EAsia = c(
    "Vietnam",
    "Laos",
    "Cambodia",
    "Thailand",
    "China",
    "Hong",
    "Taiwan",
    "Philippines",
    "Japan"
  ),
  SAsia = c("India", "Iran"),
  NorthAmerica = c("Canada", "Mexico", "United-States"),
  CentrAmerica = c(
    "Cuba",
    "Dominican-Republic",
    "Guatemala",
    "Haiti",
    "Honduras",
    "Jamaica",
    "Trinadad&Tobago",
    "Nicaragua",
    "El-Salvador",
    "",
    "?"
  ),
  SouthAmerica = c("Ecuador", "Peru", "Columbia", "South"),
  Europe = c(
    "France",
    "Germany",
    "Greece",
    "Holand-Netherlands",
    "Italy",
    "Hungary",
    "Ireland",
    "Poland",
    "Portugal",
    "Scotland",
    "England",
    "Yugoslavia"
  ),
  USTerritory = c("Outlying-US(Guam-USVI-etc)", "Puerto-Rico")
)

# Recode the native_region column using the recode function
adult$native_region <- sapply(adult$native_country, recode_value)

# Convert the recoded native_region to a factor
adult$native_region <-
  factor(adult$native_region, levels = names(new_levels))
levels(adult$native_region)

summary(adult$native_region)
adult <- adult %>% dplyr::select(-native_country)
```

### Refactor Race

```{r refactored race, warning=FALSE}
# Define the new levels for race
adult$race <- adult$race %>%
  fct_recode(
    "Native" = " Amer-Indian-Eskimo",
    "Asian_Pacific" = " Asian-Pac-Islander",
    "Black" = " Black",
    "Other" = " Other",
    "White" = " White"
  )

summary(adult$race)
```

### Remove Similiar Pop Count - Not a viable predictor
```{r}
#  Removing similar_pop_count as it does not appear to be a relevant predictor 

adult <- adult %>% dplyr::select(-similar_pop_count)
glimpse(adult)
```
## Handling Missing Data

```{r replace missing values}
# Basic descriptive statistics for numerical features
sapply(adult[sapply(adult, is.numeric)], summary)

# Summary Missing Data
sapply(adult, function(x) sum(is.na(x)))
vis_miss(adult)
         
# get the average of all remaining missing values
replace_missing_with_average <- function(data) {
  for (col_name in names(data)) {
    # Check if the column is numeric
    if (is.numeric(data[[col_name]])) {
      # Calculate the average, excluding NA values
      column_mean <- mean(data[[col_name]], na.rm = TRUE)
      # Replace NA values with the column's average
      data[[col_name]][is.na(data[[col_name]])] <- column_mean
    }
  }
  return(data)
}

# Call the replace function:
adult <- replace_missing_with_average(adult)

# Make sure categorical variables are factored
adult[] <-
  lapply(adult, function(x) {
    if (is.character(x)) {
      as.factor(x)
    } else {
      x
    }
  })

# Review changes
glimpse(adult)
```

# Section 2: EDA with Modified Adult Dataset

## Investigate the Capital Gain / Loss Data

```{r capital_gain_loss to invested, echo=FALSE}
library(ggmosaic)
library(ggthemes)

adult <- adult %>%
  mutate(invested = ifelse(capital_gain > 0 | capital_loss > 0, 1, 0))

adult$invested <- factor(adult$invested,
  levels = c("0", "1"),
  labels = c("Not_Invested", "Invested")
)

invest_df <- adult %>% dplyr::select(invested, capital_gain, capital_loss, income)
freq(data = invest_df, input = c("invested"))

# Create the basic mosaic plot
p <- ggplot(data = invest_df) +
  geom_mosaic(aes(weight = 1, x = product(invested), fill = income)) +
  theme_minimal()

# Percentage for each quadrant
counts <- invest_df %>%
  count(income, invested) %>%
  mutate(perc = n / sum(n))
print(counts)

# Mosaic plot
p <- ggplot(data = adult) +
  geom_mosaic(aes(weight = 1, x = product(invested), fill = income),
    show.legend = FALSE
  ) +
  theme_minimal()

# Calculate the counts and percentages
counts <- adult %>%
  count(income, invested) %>%
  mutate(perc = n / sum(n) * 100)

counts$label_x <- c(0.5, .95, 0.5, .95)
counts$label_y <- c(0.4, 0.25, 0.9, 0.7)

# Add the text using geom_text() and the manually estimated positions
p + geom_text(
  data = counts,
  aes(x = label_x, y = label_y, label = paste0(round(perc, 2), "%")),
  size = 4,
  color = "white"
)

ggplot(adult, aes(x = invested, fill = income)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(y = "Percentage")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))+
  ggtitle('Earnings by Investment (Percentage')

# model <- glm(income ~ ., data = adult, family = "binomial")
# ggplot(adult,(aes(x=age,y=income))) + 
#   geom_point() +
#   geom_smooth(method="loess",span=.1)


# Log-transform 'capital_gain' and 'capital_loss" for scaling purposes
adult$capital_gain_log <- log(adult$capital_gain + 1)  
adult$capital_loss_log <- log(adult$capital_loss + 1)  

# Violin Plot
ggplot(adult, aes(x = income, y = capital_gain_log, fill = income)) +   facet_wrap(~invested) +  
  geom_violin(aes(color = income), adjust = 1.5) + 
  labs(title = "Violin Plot of Capital Gain Log by Income",
       x = "Income Category",
       y = "Log of Capital Gain") 

# Boxplot  
ggplot(adult, aes(x = income, y = capital_gain_log, fill = income)) +
  geom_boxplot(aes(color = income), outlier.shape = 1) + 
  labs(title = "Boxplot of Capital Gain Log by Income",
       x = "Income Category",
       y = "Log of Capital Gain") 

# Violin Plot with invested interaction
ggplot(adult, aes(x = income, y = capital_gain_log, fill = income)) +
  geom_violin(aes(color = income), adjust = 1.5) + 
  facet_wrap(~invested) +  
  labs(title = "Violin Plot of Capital Gain Log by Income and Invested Status",
       x = "Income Category",
       y = "Log of Capital Gain") +
  theme_minimal()

# Boxplot  with invested interaction
ggplot(adult, aes(x = income, y = capital_gain_log, fill = income)) +
  geom_boxplot(aes(color = income), outlier.shape = 1) +  # Display outliers
  facet_wrap(~invested) +  # Faceting by 'invested'
  labs(title = "Boxplot of Capital Gain Log by Income and Invested Status",
       x = "Income Category",
       y = "Log of Capital Gain") +
  theme_minimal()

adult <- adult %>% dplyr::select(-capital_gain, -capital_loss)
```

```{r summary-analysis}
# Basic summary
summary(adult)

```

```{r data-visualization, warning=FALSE}
# plot_bar(adult, by = "income", nrow = 4)
# plot_boxplot(adult, by = "income")
# plot_qq(adult, by = "income")
# freq(data = adult, input = colnames)

# Bar Chart for Income
ggplot(adult, aes(x = income)) +
  geom_bar(fill = "lightblue", color = "black") +
  ggtitle("Count of Income Levels") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Calculate the count and percentage of each income level correctly
adult_percent <- adult %>%
  count(income) %>%
  mutate(perc = n / sum(n) * 100)

ggplot(adult_percent, aes(x = income, y = perc, fill = income)) +
  geom_bar(stat = "identity", color = "black") +
  scale_fill_manual(values = c("lightblue", "lightcoral")) +
  ggtitle("Percentage of Income Levels") +
  xlab("Income") +
  ylab("Percentage (%)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


# Histogram for Age
ggplot(adult, aes(x = age, fill = income)) +
  geom_histogram(binwidth = 1) +
  facet_wrap(~income) +
  ggtitle("Age Distribution by Income") +
  xlab("Age") +
  ylab("Count")

ggplot(adult, aes(x = age, fill = income)) +
  geom_histogram(aes(y = ..density.. * 100), binwidth = 1, color = "black") +
  facet_wrap(~income) +
  scale_fill_manual(values = c("<=50K" = "lightblue", ">50K" = "lightcoral")) +
  ggtitle("Age Distribution by Income (Percentage)") +
  xlab("Age") +
  ylab("Percentage (%)")


# Histogram for Gender by Income
adult_gender_income_percent <- adult %>%
  count(gender, income) %>%
  group_by(gender) %>%
  mutate(perc = n / sum(n) * 100)

ggplot(adult_gender_income_percent, aes(x = gender, y = perc, fill = income)) +
  geom_bar(stat = "identity", position = position_dodge(), color = "black") +
  scale_fill_manual(values = c("<=50K" = "lightblue", ">50K" = "lightcoral")) +
  ggtitle("Gender Distribution by Income (Percentage)") +
  xlab("Gender") +
  ylab("Percentage (%)") +
  facet_wrap(~income)


# Histogram for Hours per Week
ggplot(adult, aes(x = hours_per_week)) +
  geom_histogram(binwidth = 1, fill = "lightblue", color = "black") +
  ggtitle("Distribution of Hours per Week")



# Histogram for Education Number
ggplot(adult, aes(x = years_education, fill = income)) +
  geom_histogram(
    aes(y = ..density.. * 100),
    binwidth = 1,
    color = "black",
    position = "dodge"
  ) +
  scale_fill_manual(values = c("<=50K" = "lightblue", ">50K" = "lightcoral")) +
  ggtitle("Education Number Distribution by Income (Percentage)") +
  xlab("Education Number") +
  ylab("Percentage (%)")

# Bar Chart for Education with Ordered Levels
ggplot(adult, aes(x = education_level)) +
  geom_bar(fill = "lightblue", color = "black") +
  ggtitle("Education Level Distribution") +
  xlab("Education Level") +
  ylab("Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Bar Chart for Workclass
ggplot(adult, aes(x = workclass)) +
  geom_bar(fill = "lightblue", color = "black") +
  ggtitle("Workclass Distribution") +
  xlab("Workclass") +
  ylab("Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


# Bar Chart for Native Region
ggplot(adult, aes(x = native_region)) +
  geom_bar(fill = "lightblue", color = "black") +
  ggtitle("Count per Native Region") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Bar Chart for Relationship
ggplot(adult, aes(x = marital_status, fill = income)) +
  geom_bar(position = "dodge", color = "black") +
  scale_fill_manual(values = c("<=50K" = "lightblue", ">50K" = "lightcoral")) +
  ggtitle("Relationship Distribution by Income") +
  xlab("Relationship") +
  ylab("Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Bar Chart for Occupation
ggplot(adult, aes(x = occupation, fill = income)) +
  geom_bar(position = "dodge", color = "black") +
  scale_fill_manual(values = c("<=50K" = "lightblue", ">50K" = "lightcoral")) +
  ggtitle("Occupation Level Distribution") +
  xlab("Occupation") +
  ylab("Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Income by occupation
qplot(occupation, data = adult, fill = income) + 
  facet_grid (. ~ race) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))

# Filtering out White race so we can better visualize the other levels
race_filtered <- adult %>% filter(race != "White")

# Now create your plot with the filtered dataset
qplot(occupation, data = race_filtered, fill = income) + 
  facet_grid(. ~ race) +
  theme_clean() +  # Apply a minimal theme for a clean look
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))

ggplot(adult, aes(x = occupation, fill = income)) +
  geom_bar(position = "fill") +
  facet_grid(. ~ race) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(y = "Percentage")+
  theme_clean() +  # Apply a minimal theme for a clean look
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))

ggplot(adult, aes(x = education_level, fill = income)) +
  geom_bar(position = "fill") +
  facet_grid(. ~ gender) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(y = "Percentage")+
  theme_clean() +  # Apply a minimal theme for a clean look
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))

# adult_numeric %>% ggpairs(.,aes(colour=income))

```

## Dataset Overview

The dataset contains several variables related to demographic and employment information
of adults. Below is a brief overview of the variables included in the dataset we will be
using for modeling:

-   **Age**: Numeric. Represents the age of the individual.
-   **Workclass**: Categorical. Describes the employment status of the individual (e.g.,
    Private, Self-emp, Government).
-   **Education**: Categorical. Highest level of education achieved.
-   **Education_Num**: Numeric. Represents the total years of schooling in a continuous
    numerical form.
-   **Marital_Status**: Categorical. Marital status of the individual.
-   **Occupation**: Categorical. Type of occupation.
-   **Relationship**: Categorical. Represents the individual's role in the family.
-   **Race**: Categorical. Race of the individual.
-   **Sex**: Categorical. Gender of the individual.
-   **Capital_Gain**: (logged) Numeric. Represents income from investment sources, apart
    from wages/salary.
-   **Capital_Loss**: (logged)Numeric. Represents losses from investment sources, apart
    from wages/salary.
-   **Hours_per_Week**: Numeric. Number of hours worked per week.
-   **native_region**: Categorical. Country of origin of the individual.
-   **Income**: Categorical. Whether the individual makes more than \$50K/year.

# Section 3 Create Modeling Data and Split into Training/Testing

## Split Dataset into Train/Test

```{r create_train_test}
# Set a seed for reproducibility
set.seed(1234)

# Calculate the number of rows to select
sample_size <- round(0.8 * nrow(adult))

# Create a random sample of indices
index <- sample(1:nrow(adult), size = sample_size, replace = FALSE)

# Create the training set using the selected indices
training_set <- adult[index, ]
testing_set <- adult[-index, ]

nrow(training_set)
nrow(testing_set)

glimpse(training_set)
glimpse(testing_set)

outlier_indices <- c(31369, 15377, 20177, 06434, 32305, 3594)
training_set <- training_set[-outlier_indices, ]
```

# Generic Function to get all the statistics

```{r Print Model Metrics Function}
# Example:
# print_model_metrics(model = generic_model, test_data = testing_set, response_var = "income", pos_value = "Over.50K", threshold = 0.28, model_title = "Simple GLM")
#
library(caret)
library(pROC)

print_model_metrics <-
  function(model,
           test_data,
           response_var,
           pos_value,
           threshold = 0.5,
           model_title = "Model Metrics") {
    # Ensure that the response variable is a factor with exactly two levels
    response_factor <-
      factor(test_data[[response_var]], levels = unique(test_data[[response_var]]))
    
    # Make sure we have exactly two levels
    if (length(levels(response_factor)) != 2) {
      stop("Response variable must have exactly two levels.")
    }
    
    # Make predictions and convert to factor with the specified positive value
    probabilities <- predict(model, test_data, type = "response")
    predicted_classes <-
      factor(ifelse(
        probabilities > threshold,
        levels(response_factor)[2],
        levels(response_factor)[1]
      ),
      levels = levels(response_factor))
    
    # Calculate the confusion matrix using caret, using pos_value to specify the positive class
    cm <-
      confusionMatrix(predicted_classes, response_factor, positive = pos_value)
    
    # Extract metrics from the confusion matrix
    accuracy <- cm$overall['Accuracy']
    sensitivity <- cm$byClass['Sensitivity']
    specificity <- cm$byClass['Specificity']
    prevalence <- cm$byClass['Prevalence']
    ppv <- cm$byClass['Pos Pred Value']
    npv <- cm$byClass['Neg Pred Value']
    
    # Calculate AUROC using pROC, ensuring numeric conversion accounts for positive class being '1'
    roc_obj <-
      roc(response = as.numeric(response_factor) - 1,
          predictor = probabilities)
    auroc <- auc(roc_obj)
    
    # Replace NA values with zero or another appropriate placeholder
    metrics_values <-
      c(accuracy,
        sensitivity,
        specificity,
        prevalence,
        ppv,
        npv,
        auroc)
    metrics_values[is.na(metrics_values)] <- 0
    
    # Compile the metrics into a data frame
    metrics <- data.frame(
      Metric = c(
        "Accuracy",
        "Sensitivity",
        "Specificity",
        "Prevalence",
        "PPV",
        "NPV",
        "AUROC"
      ),
      Value = metrics_values
    )
    
    # Prepare the caption with the threshold information
    aic_val <- AIC(model)
    
    full_caption <-
      sprintf("%s\nThreshold: %s\nAIC: %.2f",
              model_title,
              threshold,
              aic_val)
    
    # Print the metrics table using knitr
    knitr::kable(
      metrics,
      col.names = c("Metric", "Value"),
      format = "markdown",
      caption = full_caption
    )
  }

```

# Generic Function to plot metrics by threshold

```{r}
library(caret)
library(pROC)
library(ggplot2)

# Example
# plot_model_metrics_by_threshold(model = simple.fit, test_data = testing_set, response_var = "income", pos_value = "Greater.50K")

plot_model_metrics_by_threshold <-
  function(model,
           test_data,
           response_var,
           pos_value,
           thresholds = seq(0.1, 0.9, by = 0.05),
           model_title = "Model Metrics by Threshold") {
    # Ensure the response variable is a factor with exactly two levels
    response_factor <-
      factor(test_data[[response_var]], levels = unique(test_data[[response_var]]))
    
    # Make sure we have exactly two levels
    if (length(levels(response_factor)) != 2) {
      stop("Response variable must have exactly two levels.")
    }
    
    # Prepare to collect metrics
    results <-
      data.frame(
        Threshold = numeric(0),
        Accuracy = numeric(0),
        Sensitivity = numeric(0),
        Specificity = numeric(0),
        Prevalence = numeric(0),
        PPV = numeric(0),
        NPV = numeric(0)
      )
    
    # Loop over each threshold
    for (threshold in thresholds) {
      # Make predictions and convert to factor with the specified positive value
      probabilities <- predict(model, test_data, type = "response")
      predicted_classes <-
        factor(
          ifelse(
            probabilities > threshold,
            pos_value,
            levels(response_factor)[levels(response_factor) != pos_value]
          ),
          levels = levels(response_factor)
        )
      
      # Calculate the confusion matrix using caret
      cm <-
        confusionMatrix(predicted_classes, response_factor, positive = pos_value)
      
      # Store metrics
      results <- rbind(
        results,
        data.frame(
          Threshold = threshold,
          Accuracy = cm$overall['Accuracy'],
          Sensitivity = cm$byClass['Sensitivity'],
          Specificity = cm$byClass['Specificity'],
          Prevalence <- cm$byClass['Prevalence'],
          PPV = cm$byClass['Pos Pred Value'],
          NPV = cm$byClass['Neg Pred Value']
        )
      )
    }
    
    # Melt results for plotting
    results_long <-
      reshape2::melt(results, id.vars = 'Threshold', variable.name = 'Metric')
    
    # Plot results
    plot <-
      ggplot(results_long, aes(x = Threshold, y = value, color = Metric)) +
      geom_line() +
      labs(title = model_title, x = "Threshold", y = "Metric Value") +
      scale_color_manual(
        values = c(
          "Accuracy" = "blue",
          "Sensitivity" = "red",
          "Specificity" = "green",
          "Prevalence" = "purple",
          "PPV" = "orange",
          "NPV" = "brown"
        )
      ) +
      scale_y_continuous(limits = c(0, 1)) +
      ggtitle(paste("Metrics for Various Thresholds -", model_title))
    
    # Return the plot
    return(plot)
  }

```

# Simple GLM model to get an idea and baseline model

```{r, warning=FALSE}

simple.fit <- glm(income ~ ., data = training_set, family = "binomial")
summary(simple.fit)

plot_model_metrics_by_threshold(
  model = simple.fit,
  test_data = testing_set,
  response_var = "income",
  pos_value = "Greater.50K",
  model_title = "Simple GLM"
)

print_model_metrics(
  model = simple.fit,
  test_data = testing_set,
  response_var = "income",
  pos_value = "Greater.50K",
  threshold = 0.25,
  model_title = "Simple GLM"
)

# AIC(simple.fit)
# conf_int <- confint(simple.fit, level = 0.95)

# Plot the ROC curve
predicted_prob <- predict(simple.fit, newdata = testing_set, type = "response")
simple.roc <- roc(testing_set$income, predicted_prob)
auc_simple <- auc(simple.roc)
plot(simple.roc, main=paste("ROC Curve (AUC =", round(auc_simple, 3), ")"))
predicted <- as.factor(ifelse(predicted_prob >= 0.28, "Greater.50K", "Less.50K"))
actual <- testing_set$income
confusion_matrix <- confusionMatrix(predicted, actual, positive="Greater.50K")
print(confusion_matrix)



```

```{r}
# 
# training_set$income.num <- ifelse(training_set$income == ">Greater.50K", 1, 0)
#     ggplot(training_set, aes(x=education_level, y=income.num)) +
#       geom_point() + 
#       geom_smooth(method="loess", formula = y ~ x, span=0.5, color="blue")
```

```{r Influential observations}
# library(car) 
influence_measures <- influence.measures(simple.fit)

# View the diagnostics
print(influence_measures)

# More detailed diagnostics
influenceIndexPlot(simple.fit)

# library(olsrr)
# ols_plot_diagnostics(simple.fit)

#Cooks D Plot
cutoff <- 4/(nrow(simple.fit)-length(simple.fit$coefficients)-2)
plot(simple.fit, which=4, cook.levels=cutoff)
abline(h=cutoff, lty=2, col="red")

#Influence Plot
influencePlot(simple.fit,
              id.method = "identify",
              main = "Influence Plot",
              sub = "Circle size is proportional to Cook’s distance")
```

## Removing outliers

```{r}
# Remove high leverage or high influence points
outlier_indices <- c(31369, 15377, 20177, 06434, 32305, 3594)
training_set <- training_set[-outlier_indices, ]
```

# Refit with removed outliers

```{r}
# Refit the model
simple.fit.clearned <- glm(income ~ ., data = training_set, family = "binomial")
summary(simple.fit.clearned)
par(mfrow=c(2,2))
plot(simple.fit.clearned)
par(mfrow=c(1,1))

plot_model_metrics_by_threshold(
  model = simple.fit.clearned,
  test_data = testing_set,
  response_var = "income",
  pos_value = "Greater.50K",
  model_title = "Simple Fit Cleaned"
)

print_model_metrics(
  model = simple.fit.clearned,
  test_data = testing_set,
  response_var = "income",
  pos_value = "Greater.50K",
  threshold = 0.267,
  model_title = "Simple Fit Cleaned"
)
```

No real improvement in metrics, so leaving the outliers in.

# Simple 2 GLM cleaned with only Significant from Simple GLM Model

```{r}
model_formula <- as.formula("income ~ age + 
                                      years_education + 
                                      marital_status + 
                                      household_role + 
                                      hours_per_week +
                                      capital_gain_log +
                                      capital_loss_log")

simple2.fit <- glm(model_formula, data = training_set, family = "binomial")
summary(simple2.fit)
par(mfrow=c(2,2))
plot(simple2.fit)
par(mfrow=c(1,1))

plot_model_metrics_by_threshold(
  model = simple2.fit,
  test_data = testing_set,
  response_var = "income",
  pos_value = "Greater.50K",
  model_title = "Simple Fit Cleaned"
)

simple <-
  print_model_metrics(
    model = simple2.fit,
    test_data = testing_set,
    response_var = "income",
    pos_value = "Greater.50K",
    threshold = 0.267,
    model_title = "Simple Fit Cleaned"
  )
print(simple)
```

$$
\hat{Y}_{\text{income}} = \beta_0 + \beta_1 \text{age} + \beta_2 \text{years_education} + \beta_3 \text{marital_status} + \beta_4 \text{household_role} + \beta_5 \text{gender} + \beta_6 \text{hours_per_week} + \beta_7 \text{capital_gain_log} + \beta_8 \text{capital_loss_log}
$$ \
# Complex Models
### Complex Poly Squared GLM cleaned with only Significant from Simple GLM Model

```{r}
model_formula <- as.formula("income ~ poly(age, 2) + 
                                      poly(years_education, 2) + 
                                      marital_status + 
                                      household_role + 
                                      poly(hours_per_week, 2) +
                                      poly(capital_gain_log, 2) +
                                      poly(capital_loss_log, 2)")


poly.fit <- glm(model_formula, data = training_set, family = "binomial")
summary(poly.fit)
par(mfrow=c(2,2))
plot(poly.fit)
par(mfrow=c(1,1))

plot_model_metrics_by_threshold(model = poly.fit, test_data = testing_set, 
                                response_var = "income", 
                                pos_value = "Greater.50K", 
                                model_title = "Simple Fit with Polynomial ^2")

poly2 <- print_model_metrics(model = poly.fit, 
                             test_data = testing_set, 
                             response_var = "income", 
                             pos_value = "Greater.50K", 
                             threshold = 0.267, 
                             model_title = "Simple Fit with Polynomial ^2")
print(poly2)
```

$$
\begin{aligned}
\hat{Y}_{\text{income}} = \beta_0 &+ \beta_{\text{age1}} \times \text{age} + \beta_{\text{age2}} \times \text{age}^2 \\
&+ \beta_{\text{ed1}} \times \text{years_education} + \beta_{\text{ed2}} \times \text{years_education}^2 \\
&+ \beta_{\text{ms}, j} \times \text{marital_status}_j \\
&+\beta_{\text{hr}, k} \times \text{household_role}_k \\
&+ \beta_{\text{hpw1}} \times \text{hours_per_week} + \beta_{\text{hpw2}} \times \text{hours_per_week}^2 \\
&+ \beta_{\text{cgl1}} \times \text{capital_gain_log} + \beta_{\text{cgl2}} \times \text{capital_gain_log}^2 \\
&+ \beta_{\text{cll1}} \times \text{capital_loss_log} + \beta_{\text{cll2}} \times \text{capital_loss_log}^2
\end{aligned}
$$

### Complex Poly Cubed GLM cleaned with only Significant from Simple GLM Model

```{r}
model_formula <- as.formula("income ~ poly(age, 3) + 
                                      poly(years_education, 3) + 
                                      marital_status + 
                                      household_role + 
                                      poly(hours_per_week, 3) +
                                      poly(capital_gain_log, 3) +
                                      poly(capital_loss_log, 3)")


poly3.fit <- glm(model_formula, data = training_set, family = "binomial")
summary(poly3.fit)
par(mfrow=c(2,2))
plot(poly3.fit)
par(mfrow=c(1,1))


plot_model_metrics_by_threshold(
  model = poly3.fit,
  test_data = testing_set,
  response_var = "income",
  pos_value = "Greater.50K",
  model_title = "Simple Fit with Polynomial ^3"
)

poly3 <-
  print_model_metrics(
    model = poly3.fit,
    test_data = testing_set,
    response_var = "income",
    pos_value = "Greater.50K",
    threshold = 0.267,
    model_title = "Simple Fit with Polynomial ^3"
  )
print(poly3)
```


### Complex Model interaction Gender

```{r, warning=FALSE}
model_formula <- as.formula("income ~ (age + 
                                      years_education + 
                                      marital_status + 
                                      household_role + 
                                      hours_per_week +
                                      capital_gain_log +
                                      capital_loss_log) * gender")

complex.fit <- glm(model_formula, data = training_set, family = "binomial")
summary(complex.fit)
par(mfrow=c(2,2))
plot(complex.fit)
par(mfrow=c(1,1))

plot_model_metrics_by_threshold(
  model = complex.fit,
  test_data = testing_set,
  response_var = "income",
  pos_value = "Greater.50K",
  model_title = "Complex (Interaction Gender)"
)

complex0 <-
  print_model_metrics(
    model = complex.fit,
    test_data = testing_set,
    response_var = "income",
    pos_value = "Greater.50K",
    threshold = 0.267,
    model_title = "Complex (Interaction Gender)"
  )
print(complex0)

```

```{r}
model_formula <- as.formula("income ~ (age + 
                                      years_education + 
                                      marital_status + 
                                      household_role + 
                                      hours_per_week +
                                      poly(capital_gain_log,2) +
                                      poly(capital_loss_log,2)) * gender")

complex.fit <- glm(model_formula, data = training_set, family = "binomial")
summary(complex.fit)
par(mfrow=c(2,2))
plot(complex.fit)
par(mfrow=c(1,1))

plot_model_metrics_by_threshold(
  model = complex.fit,
  test_data = testing_set,
  response_var = "income",
  pos_value = "Greater.50K",
  model_title = "Complex (Interaction Gender with poly)"
)

complex1 <-
  print_model_metrics(
    model = complex.fit,
    test_data = testing_set,
    response_var = "income",
    pos_value = "Greater.50K",
    threshold = 0.267,
    model_title = "Complex (Interaction Gender with poly)"
  )
print(complex1)
```

### Complex Model interaction Native Region

```{r}
# Using Native Country as interaction except for capital_gain_log and capital_loss_log which is using invested
model_formula <- as.formula("income ~ (age + 
                                      years_education + 
                                      marital_status + 
                                      household_role + 
                                      hours_per_week +
                                      capital_gain_log +
                                      capital_loss_log) * native_region")


complex.fit <- glm(model_formula, data = training_set, family = "binomial")
summary(complex.fit)
par(mfrow=c(2,2))
plot(complex.fit)
par(mfrow=c(1,1))

plot_model_metrics_by_threshold(
  model = complex.fit,
  test_data = testing_set,
  response_var = "income",
  pos_value = "Greater.50K",
  model_title = "Complex (Interaction Native Region)"
)

complex2 <-
  print_model_metrics(
    model = complex.fit,
    test_data = testing_set,
    response_var = "income",
    pos_value = "Greater.50K",
    threshold = 0.267,
    model_title = "Complex (Interaction Native Region)"
  )
print(complex2)
```
$$
\hat{Y}_{\text{income}} = \beta_0 + (\beta_{\text{age}} \times \text{age} + \beta_{\text{age,nr}} \times \text{age} \times \text{native_region}) + \\
(\beta_{\text{ye}} \times \text{years_education} + \beta_{\text{ye,nr}} \times \text{years_education} \times \text{native_region}) + \\
(\beta_{\text{ms}, j} \times \text{marital_status}_j + \beta_{\text{ms,nr}, j} \times \text{marital_status}_j \times \text{native_region}) + \\
(\beta_{\text{hr}, k} \times \text{household_role}_k + \beta_{\text{hr,nr}, k} \times \text{household_role}_k \times \text{native_region}) + \\
(\beta_{\text{hpw}} \times \text{hours_per_week} + \beta_{\text{hpw,nr}} \times \text{hours_per_week} \times \text{native_region}) + \\
(\beta_{\text{cgl}} \times \text{capital_gain_log} + \beta_{\text{cgl,nr}} \times \text{capital_gain_log} \times \text{native_region}) + \\
(\beta_{\text{cll}} \times \text{capital_loss_log} + \beta_{\text{cll,nr}} \times \text{capital_loss_log} \times \text{native_region})

$$


### Complex Model interaction Native Region with Poly

```{r}
# Using Native Country as interaction except for capital_gain_log and capital_loss_log which is using invested
model_formula <- as.formula("income ~ (age + 
                                      years_education + 
                                      marital_status + 
                                      household_role + 
                                      hours_per_week +
                                      poly(capital_gain_log,2) +
                                      poly(capital_loss_log,2)) * native_region")


complex.fit <- glm(model_formula, data = training_set, family = "binomial")
summary(complex.fit)
par(mfrow=c(2,2))
plot(complex.fit)
par(mfrow=c(1,1))

plot_model_metrics_by_threshold(
  model = complex.fit,
  test_data = testing_set,
  response_var = "income",
  pos_value = "Greater.50K",
  model_title = "Complex (Interaction Native Region with poly^2)"
)

complex2poly <-
  print_model_metrics(
    model = complex.fit,
    test_data = testing_set,
    response_var = "income",
    pos_value = "Greater.50K",
    threshold = 0.267,
    model_title = "Complex (Interaction Native Region with poly^2)"
  )
print(complex2poly)
```
$$
\hat{Y}_{\text{income}} = \beta_0 + (\beta_{\text{age}} \times \text{age} + \beta_{\text{age}^2} \times \text{age}^2 + \beta_{\text{age,nr}} \times \text{age} \times \text{native_region} + \beta_{\text{age}^2\text{,nr}} \times \text{age}^2 \times \text{native_region}) + \\
(\beta_{\text{ye}} \times \text{years_education} + \beta_{\text{ye}^2} \times \text{years_education}^2 + \beta_{\text{ye,nr}} \times \text{years_education} \times \text{native_region} + \beta_{\text{ye}^2\text{,nr}} \times \text{years_education}^2 \times \text{native_region}) + \\
(\beta_{\text{ms}, j} \times \text{marital_status}_j + \beta_{\text{ms,nr}, j} \times \text{marital_status}_j \times \text{native_region}) + \\
(\beta_{\text{hr}, k} \times \text{household_role}_k + \beta_{\text{hr,nr}, k} \times \text{household_role}_k \times \text{native_region}) + \\
(\beta_{\text{hpw}} \times \text{hours_per_week} + \beta_{\text{hpw}^2} \times \text{hours_per_week}^2 + \beta_{\text{hpw,nr}} \times \text{hours_per_week} \times \text{native_region} + \beta_{\text{hpw}^2\text{,nr}} \times \text{hours_per_week}^2 \times \text{native_region}) + \\
(\beta_{\text{cgl1}} \times \text{capital_gain_log} + \beta_{\text{cgl2}} \times \text{capital_gain_log}^2 + \beta_{\text{cgl,nr}} \times \text{capital_gain_log} \times \text{native_region} + \beta_{\text{cgl2,nr}} \times \text{capital_gain_log}^2 \times \text{native_region}) + \\
(\beta_{\text{cll1}} \times \text{capital_loss_log} + \beta_{\text{cll2}} \times \text{capital_loss_log}^2 + \beta_{\text{cll,nr}} \times \text{capital_loss_log} \times \text{native_region} + \beta_{\text{cll2,nr}} \times \text{capital_loss_log}^2 \times \text{native_region})
$$

### Complex Model interaction Race

```{r}
# Using Race as interaction except for capital_gain_log and capital_loss_log which is using invested
model_formula <- as.formula("income ~ (age + 
                                      years_education + 
                                      marital_status + 
                                      household_role + 
                                      hours_per_week +
                                      poly(capital_gain_log, 2)+
                                      poly(capital_loss_log, 2)) * race")


complex.fit <- glm(model_formula, data = training_set, family = "binomial")
summary(complex.fit)
par(mfrow=c(2,2))
plot(complex.fit)
par(mfrow=c(1,1))

plot_model_metrics_by_threshold(
  model = complex.fit,
  test_data = testing_set,
  response_var = "income",
  pos_value = "Greater.50K",
  model_title = "Complex (Interaction Race)"
)

complex3 <-
  print_model_metrics(
    model = complex.fit,
    test_data = testing_set,
    response_var = "income",
    pos_value = "Greater.50K",
    threshold = 0.267,
    model_title = "Complex (Interaction Race)"
  )
print(complex3)
```

### Complex Model interaction Occupation

```{r, warning=FALSE}
# Using Race as interaction except for capital_gain_log and capital_loss_log which is using invested
model_formula <- as.formula("income ~ (age +
                                      years_education +
                                      hours_per_week +
                                      poly(capital_gain_log, 2)+
                                      poly(capital_loss_log, 2)) * occupation")


complex.fit <- glm(model_formula, data = training_set, family = "binomial")
summary(complex.fit)
par(mfrow=c(2,2))
plot(complex.fit)
par(mfrow=c(1,1))

plot_model_metrics_by_threshold(
  model = complex.fit,
  test_data = testing_set,
  response_var = "income",
  pos_value = "Greater.50K",
  model_title = "Complex (Interaction Occupation)"
)

complex4 <-
  print_model_metrics(
    model = complex.fit,
    test_data = testing_set,
    response_var = "income",
    pos_value = "Greater.50K",
    threshold = 0.267,
    model_title = "Complex (Interaction Occupation)"
  )
print(complex4)
```

$$
\hat{Y}_{\text{income}} = \beta_0 + (\beta_1 \times \text{age} \times \text{occupation}) + (\beta_2 \times \text{years_education} \times \text{occupation}) + (\beta_3 \times \text{marital_status} \times \text{occupation}) + (\beta_4 \times \text{household_role} \times \text{occupation}) + (\beta_5 \times \text{hours_per_week} \times \text{occupation}) \\
+ (\beta_6 \times \text{capital_gain_log} \times \text{occupation}) + (\beta_7 \times \text{capital_loss_log} \times \text{occupation})
$$

### Complex Model interaction Household Role 

```{r, warning=FALSE}
model_formula <- as.formula("income ~ (age +
                                      years_education +
                                      marital_status +
                                      hours_per_week +
                                      capital_gain_log +
                                      capital_loss_log) * household_role")


complex.fit <- glm(model_formula, data = training_set, family = "binomial")
summary(complex.fit)
par(mfrow=c(2,2))
plot(complex.fit)
par(mfrow=c(1,1))

plot_model_metrics_by_threshold(
  model = complex.fit,
  test_data = testing_set,
  response_var = "income",
  pos_value = "Greater.50K",
  model_title = "Complex (Interaction Household Role)"
)

complex5 <-
  print_model_metrics(
    model = complex.fit,
    test_data = testing_set,
    response_var = "income",
    pos_value = "Greater.50K",
    threshold = 0.267,
    model_title = "Complex (Interaction Household Role)"
  )

print(complex5)
```

### Complex Model interaction Household Role with poly

```{r, warning=FALSE}

model_formula <- as.formula("income ~ (age +
                                    years_education +
                                    hours_per_week +
                                    poly(capital_gain_log, 2)+
                                    poly(capital_loss_log, 2)) * household_role")


complex.fit <- glm(model_formula, data = training_set, family = "binomial")
summary(complex.fit)
par(mfrow=c(2,2))
plot(complex.fit)
par(mfrow=c(1,1))

plot_model_metrics_by_threshold(
  model = complex.fit,
  test_data = testing_set,
  response_var = "income",
  pos_value = "Greater.50K",
  model_title = "Complex (Interaction Household Role and polynomial)"
)

complex6 <-
  print_model_metrics(
    model = complex.fit,
    test_data = testing_set,
    response_var = "income",
    pos_value = "Greater.50K",
    threshold = 0.267,
    model_title = "Complex (Interaction Household Role and polynomial)"
  )

print(complex6)
```



## Print all complex together

```{r, warning=FALSE, echo=FALSE}
# library(gridExtra)
# grid.table(list(complex1, complex2, complex3, complex4))
print(simple)
print(poly2)
print(poly3)
print(complex0)
print(complex1)
print(complex2)
print(complex2poly)
print(complex3)
print(complex4)
print(complex5)
print(complex6)
```

# Feature Selections using AIC as the criterion

### Forward Selection

```{r}
library(stats)

# Start with a null model with no predictors, just an intercept
null_model <- glm(income ~ 1, family = binomial(), data = training_set)

# Full model with all predictors
full_model <- glm(income ~ ., family = binomial(), data = training_set)


forward_model <- step(null_model, scope = list(lower = null_model, 
                                                   upper = full_model), 
                          direction = "forward", trace = 2)

summary(forward_model)
par(mfrow=c(2,2))
plot(forward_model)
par(mfrow=c(1,1))

best_foward <- formula(forward_model)
print(best_foward)

```

### Backward Selection

```{r}
full_model <- glm(income ~ ., family = binomial(), data = training_set)

backward_model <- step(full_model, direction = "backward", trace = 2)

summary(backward_model)
par(mfrow=c(2,2))
plot(backward_model)
par(mfrow=c(1,1))

best_backward <- formula(backward_model)
print(best_backward)

```

### Stepwise Selection

```{r}
full_model <- glm(income ~ ., family = binomial(), data = training_set)

stepwise_model <- step(full_model, direction = "both", trace = 2)

summary(stepwise_model)
par(mfrow=c(2,2))
plot(stepwise_model)
par(mfrow=c(1,1))

best_both <- formula(stepwise_model)
print(best_both)

```

```{r}
print(best_foward)
print(best_backward)
print(best_both)
```

# Feature Forward model with only significant predictors

```{r, warning=FALSE}

model_formula <- as.formula(best_foward)

model_forward <- glm(model_formula, data = training_set, family = "binomial")
summary(model_forward)

plot_model_metrics_by_threshold(model = model_forward, test_data = testing_set, response_var = "income", pos_value = "Greater.50K", model_title = "Feature Forward")

print_model_metrics(model = model_forward, test_data = testing_set, response_var = "income", pos_value = "Greater.50K", threshold = 0.26, model_title = "Feature Forward")

```

# Feature Backward model with only significant predictors

```{r, warning=FALSE}

model_formula <- as.formula(best_backward)

model_backward <- glm(model_formula, data = training_set, family = "binomial")
summary(model_backward)

plot_model_metrics_by_threshold(model = model_backward, test_data = testing_set, response_var = "income", pos_value = "Greater.50K", model_title = "Feature Backward")

print_model_metrics(model = model_backward, test_data = testing_set, response_var = "income", pos_value = "Greater.50K", threshold = 0.26, model_title = "Feature Backward")

```


### PCA with just numerics

```{r}
library(stats)
library(caret)

# Create a dummy variable model
train_numeric <- dplyr::select(training_set, dplyr::where(is.numeric))
glimpse(train_numeric)

pca_result <- prcomp(train_numeric, center = TRUE, scale. = TRUE)
summary(pca_result)

loadings <- pca_result$rotation
print(loadings)
print("Coefficients - PC1:")
print(loadings[,1])

cumulative_variance <- cumsum(pca_result$sdev^2 / sum(pca_result$sdev^2))
num_pcs <- which(cumulative_variance >= 0.8)[1]

plot(cumulative_variance, xlab = "Number of Components", ylab = "Cumulative Proportion of Variance Explained", type = 'b', 
     main = "Scree Plot")
abline(v = num_pcs, col = "blue", lty = 2)
abline(h = cumulative_variance[num_pcs], col = "red", lty = 2)
legend("bottomright", legend = paste("80% variance at", num_pcs, "PCs"), bty = "n")
print(paste("Number of PCs:", num_pcs))
print(paste("Cumulative Variance Explained:", round(cumulative_variance[num_pcs] * 100, 2), "%"))
```

### PCA with dummies

```{r PCA Dummy}
# library(stats)
# library(caret)
# 
# # Create a dummy variable model
# train_subset <- training_set %>% dplyr::select(-income)
# dummied_model <- dummyVars(~ ., data = train_subset)
# training_transformed <- predict(dummied_model, newdata = train_subset)
# train_numeric <- data.frame(training_transformed)
# head(train_numeric)
# 
# train_numeric <- train_numeric %>% dplyr::select(-income)
# pca_result <- prcomp(train_numeric, center = TRUE, scale. = TRUE)
# summary(pca_result)
# 
# loadings <- pca_result$rotation
# print(loadings)
# print("Coefficients - PC1:")
# print(loadings[,1])
# 
# cumulative_variance <- cumsum(pca_result$sdev^2 / sum(pca_result$sdev^2))
# num_pcs <- which(cumulative_variance >= 0.8)[1]
# 
# plot(cumulative_variance, xlab = "Number of Components", ylab = "Cumulative Proportion of Variance Explained", type = 'b', 
#      main = "Scree Plot")
# abline(v = num_pcs, col = "blue", lty = 2)
# abline(h = cumulative_variance[num_pcs], col = "red", lty = 2)
# legend("bottomright", legend = paste("80% variance at", num_pcs, "PCs"), bty = "n")
# print(paste("Number of PCs:", num_pcs))
# print(paste("Cumulative Variance Explained:", round(cumulative_variance[num_pcs] * 100, 2), "%"))
# 
```

#### PCA Scatterplots

```{r}
pca_df <- as.data.frame(pca_result$x[, 1:5])
pca_df$income <- training_set$income

ggplot(pca_df, aes(x = PC1, y = PC2, color = income)) +
  geom_jitter() +
  theme_minimal()

ggplot(pca_df, aes(x = PC1, y = PC3, color = income)) +
  geom_jitter() +
  theme_minimal()

ggplot(pca_df, aes(x = PC1, y = PC4, color = income)) +
  geom_jitter() +
  theme_minimal()
```

#### PCA 3D Model

```{r PCA}

library(plotly)

# Create a 3D scatter plot
plot_ly(data = pca_df, x = ~PC1, y = ~PC2, z = ~PC3, 
        color = ~income, type = 'scatter3d', mode = 'markers') %>%
  layout(title = "PCA: First Three Principal Components by Income", 
         scene = list(xaxis = list(title = 'PC1'),
                      yaxis = list(title = 'PC2'),
                      zaxis = list(title = 'PC3')))
```

### Just for fun, modeling PCA with only numerics, no dummies

```{r pca_model, warning=FALSE}
# Selecting numeric columns from training and testing sets
train_numeric <-  dplyr::select(training_set, dplyr::where(is.numeric))
test_numeric <-  dplyr::select(testing_set, dplyr::where(is.numeric))
glimpse(train_numeric)

# PCA on the training data
pca_result <- prcomp(train_numeric, center = TRUE, scale. = TRUE)

# Scale the testing data using the same parameters as the training set
test_numeric_scaled <- scale(test_numeric, center = pca_result$center, scale = pca_result$scale)

# Projecting the testing data into the PCA space
test_pca <- test_numeric_scaled %*% pca_result$rotation

# Summary of the PCA on the training data
summary(pca_result)

# Create data frame from PCA results for training data
pca_df <- as.data.frame(pca_result$x[, 1:5])
pca_df$income <- training_set$income

# Create a PCA data frame for the testing data
test_pca_df <- as.data.frame(test_pca[, 1:5])
test_pca_df$income <- testing_set$income

# Model fitting using the PCA components
pca.fit <- glm(income ~ PC1 + PC2 + PC3 + PC4, data = pca_df, family = "binomial")
summary(pca.fit)

plot_model_metrics_by_threshold(model = pca.fit, test_data = test_pca_df, response_var = "income", pos_value = "Greater.50K")

print_model_metrics(model = pca.fit, test_data = test_pca_df, response_var = "income", pos_value = "Greater.50K", threshold = 0.22, model_title = "Feature GLM")

```

### LDA

```{r LDA}

library(MASS) 

fitControl<-trainControl(method="repeatedcv",number=5,repeats=1,classProbs=TRUE, summaryFunction=mnLogLoss)
lda.fit<-train(income ~ age + years_education + hours_per_week+ capital_gain_log + capital_loss_log, data=training_set,
               method="lda",
               trControl=fitControl,
               metric="logLoss")
predictions <- predict(lda.fit, testing_set, type = "raw")
head(predictions)

confusionMatrix(data = predictions, reference = testing_set$income, positive = "Greater.50K")
summary(lda.fit)
```

### QDA

```{r QDA}

library(caret)
testing_set$income <- as.factor(testing_set$income)
training_set$income <- as.factor(training_set$income)

fitControl <- trainControl(
  method = "repeatedcv",
  number = 5,
  repeats = 1,
  classProbs = TRUE,  
  summaryFunction = twoClassSummary  
)
qda.fit<-train(income ~ age + years_education + hours_per_week+ capital_gain_log + capital_loss_log, data=training_set,
               method="qda",
               trControl=fitControl,
               metric="logLoss")

predictions <- predict(qda.fit, testing_set, type = "prob")
predicted <- ifelse(predictions[,"Greater.50K"] >= 0.28, "Greater.50K", "Less.50K")
predicted <- factor(predicted, levels = levels(testing_set$income))
actual <- testing_set$income
confusion_matrix <- confusionMatrix(predicted, actual, positive = "Greater.50K")
print(confusion_matrix)

```

There is strong evidence strong evidence at the 95% confidence level that high-skill workers are approximately 4.43 times more likely to earn over $50K compared to agricultural workers. With 95% confidence that the true increase is between  3.45 to 5.70 keeping all other variables constant.

4.43