---
title: "Project 2 Predicting Salary"
author: "Adam E."
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load_packages, echo=FALSE}
library(tidyverse)
library(dplyr)
library(plyr)
library(tidyr)
library(lmtest) # Durban-Watson test for independence
library(car)
library(corrplot)
library(jtools)
library(sjPlot)
library(ResourceSelection)
library(ROCR)
library(pROC)
library(ggplot2)
library(naniar)  #for visualizing missing data
library(ResourceSelection) # Hoslem test for goodness of fit
```
# Key question for this study: Is it possible to predict when someone is likely to make over 50k? 
* Secondary question: What variables best predict a subject making over 50k?


# Data Descriptions:
* Sources: https://archive.ics.uci.edu/dataset/2/adult

## Variable Name	Role	Type	Demographic	Description	Units	Missing Values
* age	Feature	Integer	Age	N/A		no

* workclass	Feature	Categorical	Income	Private, Self-emp-not-inc, Self-emp-inc, Federal-gov, Local-gov, State-gov, Without-pay, Never-worked.		yes

* fnlwgt	Feature	Integer				no

* education	Feature	Categorical	Education Level	Bachelors, Some-college, 11th, HS-grad, Prof-school, Assoc-acdm, Assoc-voc, 9th, 7th-8th, 12th, Masters, 1st-4th, 10th, Doctorate, 5th-6th, Preschool.		no
* education-num	Feature	Integer	Education Level			no

* marital-status	Feature	Categorical	Other	Married-civ-spouse, Divorced, Never-married, Separated, Widowed, Married-spouse-absent, Married-AF-spouse.		no

* occupation	Feature	Categorical	Other	Tech-support, Craft-repair, Other-service, Sales, Exec-managerial, Prof-specialty, Handlers-cleaners, Machine-op-inspct, Adm-clerical, Farming-fishing, Transport-moving, Priv-house-serv, Protective-serv, Armed-Forces.		yes

* relationship	Feature	Categorical	Other	Wife, Own-child, Husband, Not-in-family, Other-relative, Unmarried.		no

* race	Feature	Categorical	Race	White, Asian-Pac-Islander, Amer-Indian-Eskimo, Other, Black.		no

* sex	Feature	Binary	Sex	Female, Male.		no

* capital-gain	Feature	Integer				no

* capital-loss	Feature	Integer				no

* hours-per-week	Feature	Integer				no

* native-country	Feature	Categorical	Other	 [1] ""                            " ?"                          " Cambodia"                   " Canada"                    
 [5] " China"                      " Columbia"                   " Cuba"                       " Dominican-Republic"        
 [9] " Ecuador"                    " El-Salvador"                " England"                    " France"                    
[13] " Germany"                    " Greece"                     " Guatemala"                  " Haiti"                     
[17] " Holand-Netherlands"         " Honduras"                   " Hong"                       " Hungary"                   
[21] " India"                      " Iran"                       " Ireland"                    " Italy"                     
[25] " Jamaica"                    " Japan"                      " Laos"                       " Mexico"                    
[29] " Nicaragua"                  " Outlying-US(Guam-USVI-etc)" " Peru"                       " Philippines"               
[33] " Poland"                     " Portugal"                   " Puerto-Rico"                " Scotland"                  
[37] " South"                      " Taiwan"                     " Thailand"                   " Trinadad&Tobago"           
[41] " United-States"              " Vietnam"                    " Yugoslavia"                		yes
income	Target	Binary	Income	>50K, <=50K.		no

Note: 
* fnlwgt (third column) - change to 'similar_pop_count' represents the number of subjects in the target population that the responding subject represents.
* education_num - change to 'years_eduction' - count of total years in school
* sex - changed to 'gender'
```{r pressure, echo=FALSE}
# setwd("D:/University/SMU/Applied Statistics/Project2")
adult<-read.csv("50_or_More.csv",stringsAsFactors = T)

colnames(adult) <- c('age', 'workclass', 'similar_pop_count', 'education_level', 
                     'years_education', 'marital_status', 'occupation', 'household_role', 'race', 'gender', 
                     'capital_gain', 'capital_loss', 'hours_per_week', 'native_country', 'income')

adult
summary(adult)
```


# Tidying Data and resolving missing values:
*  First get a card of missing values per column and then count them. 
**   Findings: Of 32562 data values 15 are missing (less than 0.1 percent missing).
```{r Visualize_missing_values}

#Checking data types
str(adult)
cat("\n------------------Count NA----------------------\n\n")
sort(colSums(is.na(adult)), decreasing = TRUE)
vis_miss(adult)
dim(adult)
```

# Most of the variables are categorical. To Better analyze them, we will:
  1. Reclassified values into value groups to reduce levels ahead of any EDA and predictions.
  2. replaced the remaining 15 missing values with averages.
```{r reduce_levels}
# Function to recode the values with trimmed spaces
recode_value <- function(old_level) {
  old_level <- trimws(old_level) # Trim leading and trailing whitespace
  for (category in names(new_levels)) {
    if (old_level %in% new_levels[[category]]) {
      return(category)
    }
  }
  return(NA) # Return NA if the education level does not match any category
}

levels(adult$workclass)<- c("unemployed", "unemployed","FedGov","LocGov","NeverWorked","Private",
                                   "SelfEmpInc","SelfEmpNotInc","StateGov","unemployed")


# The new education level categories
new_levels <- list(
  no_edu = c("", "Preschool"),
  primary = c("1st-4th", "5th-6th"),
  secondary = c("7th-8th"),
  highsch = c("9th", "10th", "11th", "12th", "HS-grad"),
  assoc = c("Assoc-acdm", "Assoc-voc", "Some-college"),
  undergrad = c("Bachelors", "Prof-school"),
  master = c("Masters"),
  phd = c("Doctorate")
)

# Recode the education levels with the updated function
adult$education_level <- sapply(adult$education_level, recode_value)

# recreate the levels
levels(adult$education_level)<- c("no_edu", "primary", "secondary", "highsch", "assoc", "undergrad", "master", "phd")

# The new marital status categories
new_levels <- list(
  divorce = c("Divorced", "Separated"),
  married = c("Married-AF-spouse", "Married-civ-spouse", "Married-spouse-absent"),
  notmarried = c("", "Never-married"),
  widowed = c("Widowed")
)

# Recode the marital statuses with the updated function
adult$marital_status <- sapply(adult$marital_status, recode_value)

# recreate the levels
levels(adult$marital_status)<- c("divorce", "married", "notmarried", "widowed")


new_levels <- list(
  other = c("", "?"),
  clerical = c("Adm-clerical"),
  midskill = c("Craft-repair", "Machine-op-inspct", "Transport-moving"),
  lowskill = c("Handlers-cleaners", "Other-service", "Priv-house-serv", "Armed-Forces"),
  highskill = c("Sales", "Tech-support", "Protective-serv", "Prof-specialty", "Exec-managerial"),
  agriculture = c("Farming-fishing")
)


# Recode the marital statuses with the updated function
adult$occupation <- sapply(adult$occupation, recode_value)

# recreate the levels
levels(adult$occupation)<- c("other", "clerical", "midskill", "lowskill", "highskill", "agriculture")


# Define the new levels for household_role
new_levels <- list(
  husband = c("Husband"),
  wife = c("Wife"),
  tenant = c("Not-in-family"),
  single_occ = c("Unmarried", ""),
  relative = c("Other-relative"),
  depchild = c("Own-child")
)

# Recode the household_role column using the recode function
adult$household_role <- sapply(adult$household_role, recode_value)

# Convert the recoded household_role to a factor
adult$household_role <- factor(adult$household_role, levels = names(new_levels))

# Define the new levels for native_country
new_levels <- list(
  EAsia = c("Vietnam", "Laos", "Cambodia", "Thailand", "China", "Hong", "Taiwan", "Philippines", "Japan"),
  SAsia = c("India", "Iran"),
  NorthAmerica = c("Canada", "Mexico", "United-States"),
  CentrAmerica = c("Cuba", "Dominican-Republic", "Guatemala", "Haiti", "Honduras", "Jamaica", "Trinadad&Tobago", "Nicaragua", "El-Salvador", "", "?"),
  SouthAmerica = c("Ecuador", "Peru", "Columbia", "South"),
  Europe = c("France", "Germany", "Greece", "Holand-Netherlands", "Italy", "Hungary", "Ireland", "Poland", "Portugal", "Scotland", "England", "Yugoslavia"),
   USTerritory = c("Outlying-US(Guam-USVI-etc)", "Puerto-Rico")
)

# Recode the native_country column using the recode function
adult$native_country <- sapply(adult$native_country, recode_value)

# Convert the recoded native_country to a factor
adult$native_country <- factor(adult$native_country, levels = names(new_levels))
levels(adult$native_country)


# Define the new levels for income
new_levels <- list(
  "<=50K" = c("<=50K", ""),
  ">50K" = c(">50K")
)

# Recode the income column using the recode function
adult$income <- sapply(adult$income, recode_value)

# Convert the recoded income to a factor
adult$income <- factor(adult$income, levels = names(new_levels))
levels(adult$income)

# Define the new levels for gender
new_levels <- list(
  "female" = c("Female", ""),
  "male" = c("Male")
)

# Recode the gender column using the recode function
adult$gender <- sapply(adult$gender, recode_value)

# Convert the recoded gender to a factor
adult$gender <- factor(adult$gender, levels = names(new_levels))
levels(adult$gender)

# Define the new levels for race
new_levels <- list(
  other = c("Other", ""),
  white = c("White"),
  black = c("Black"),
  native = c("Amer-Indian-Eskimo"),
  Asian_Pac_Islander = c("Asian-Pac-Islander")
)

# Recode the race column using the recode function
adult$race <- sapply(adult$race, recode_value)

# Convert the recoded race to a factor
adult$race <- factor(adult$race, levels = names(new_levels))
levels(adult$race)

# get the average of all remaining missing values
replace_missing_with_average <- function(data) {
  for (col_name in names(data)) {
    # Check if the column is numeric
    if (is.numeric(data[[col_name]])) {
      # Calculate the average, excluding NA values
      column_mean <- mean(data[[col_name]], na.rm = TRUE)
      # Replace NA values with the column's average
      data[[col_name]][is.na(data[[col_name]])] <- column_mean
    }
  }
  return(data)
}

# Call the replace function:
adult <- replace_missing_with_average(adult)

# Make sure categorical variables are factored
adult[] <- lapply(adult, function(x) if(is.character(x)) as.factor(x) else x)


# Create the spans in a new column 
adult <- adult %>%
  mutate(hours_worked_span = case_when(
    `hours_per_week` >= 1 & `hours_per_week` <= 15 ~ "1-15",
    `hours_per_week` >= 16 & `hours_per_week` <= 30 ~ "16-30",
    `hours_per_week` >= 31 & `hours_per_week` <= 45 ~ "31-45",
    `hours_per_week` >= 46 & `hours_per_week` <= 60 ~ "46-60",
    `hours_per_week` >= 61 & `hours_per_week` <= 75 ~ "61-75",
    `hours_per_week` >= 76 & `hours_per_week` <= 100 ~ "76-100",
    TRUE ~ NA_character_  # Handle any other cases
  ))

# Convert the recoded hours_worked_span to a factor
adult$hours_worked_span <- factor(adult$hours_worked_span, levels = c("1-15", "16-30", "31-45", "46-60", "61-75", "76-100"))

# Check the levels 
levels(adult$hours_worked_span)


# Create new spans and a column
adult <- adult %>%
  mutate(age_span = cut(age, breaks = c(0, 20, 35, 50, 65, 75, 100),
                        labels = c("1-20", "21-35", "36-50", "51-65", "66-75", "76-100"),
                        right = FALSE))

# Convert the 'age_span' column to a categorical type with ordered levels
adult$age_span <- factor(adult$age_span,
                         levels = c("1-20", "21-35", "36-50", "51-65", "66-75", "76-100"),
                         ordered = TRUE)

# Check the levels of the new factor
levels(adult$age_span)

# Review changes
adult

```


# Verify all missing values are adjusted.
```{r recheck_missing_values}
adult <- adult %>%
  mutate(invested = ifelse(capital_gain > 0 | capital_loss > 0, 1, 0))

adult$invested <- factor(adult$invested,
  levels = c("0", "1"),
  labels = c("Not_Invested", "Invested")
)
levels(adult$invested)

# Get all unique values in'invested'
unique_values <- unique(adult$invested)
unique_values


#Checking data types
str(adult)
cat("\n------------------Count NA----------------------\n\n")
sort(colSums(is.na(adult)), decreasing = TRUE)
vis_miss(adult)
dim(adult)
```


# EDA of the variables
Sources: https://oakleyj.github.io/MAS61004/eda-for-logistic-regression.html
```{r eda_continuous}
# histogram of age by gender group
ggplot(adult) + aes(x=age, group=income, fill=income) + 
  geom_histogram(binwidth=1, color='black') +
  labs(title = "Income by Age", x = "Age")

ggplot(adult) + aes(x=years_education, group=income, fill=income) + 
  geom_histogram(binwidth=1, color='black') +
  labs(title = "Income by Years of Education", x = "Years")

ggplot(adult) + aes(x=hours_per_week, group=income, fill=income) + 
  geom_histogram(binwidth=1, color='black') +
  labs(title = "Income by Hours Per Week", x = "Hours")
```

```{r eda_capital_Gain_loss}
# Income by gain
qplot(capital_gain, data = adult, fill = income)+
  ggtitle('Earnings with Capital Gains')

# Income by loss
qplot(capital_loss, data = adult, fill = income)+
  ggtitle('Earnings with Capital Loss')



ggplot(adult, aes(x = invested, y = capital_gain, colour = income, fill = income)) +
  geom_boxplot(color='black') +
  ggtitle("Earnings by Investment Gain")

ggplot(adult, aes(x = invested, y = capital_loss, colour = income, fill = income)) +
  geom_boxplot(color='black') +
  ggtitle("Earnings by Investment Loss")

# Income by hours with percentage
ggplot(adult, aes(x = hours_per_week, fill = income)) +
  geom_bar(position = "fill") +
  # facet_grid(. ~ race) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(y = "Percentage")+
  ggtitle("Earnings by Hours") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))

```



```{r loess_continuous}
library(tidyverse)
library(broom)
# Check the continuous variables for linearity
ggplot(adult_dum,(aes(x=age,y=income)))+geom_point()+geom_smooth(method="loess",span=.1) +
  labs(title = "Income by Age", x = "Age", y = "Income")
ggplot(adult_dum,(aes(x=years_education,y=income)))+geom_point()+geom_smooth(method="loess",span=.6)+
  labs(title = "Income by Years of Education", x = "Years", y = "Income")
ggplot(adult_dum,(aes(x=hours_per_week,y=income)))+geom_point()+geom_smooth(method="loess",span=.6)+
  labs(title = "Income by Hours Per Week", x = "Hours", y = "Income")

# Non logged capital Gains
ggplot(adult, aes(x=capital_gain, y=adult_dum$income)) + 
  geom_point() +
  geom_smooth(method="loess", size=1, span=1) +
  ylim(-.2, 2.5) +
  labs(title = "Income by Capital Gain", x = "Capital Gain", y = "Income")

# Non-logged capital loss
ggplot(adult, aes(x=capital_loss, y=adult_dum$income)) + 
  geom_point() +
  geom_smooth(method="loess", size=1, span=1) +
  ylim(-.2, 2.5) +
  labs(title = "Income by Capital Loss", x = "Capital Loss", y = "Income")


# Logged gain and loss
ggplot(adult_dum,(aes(x=capital_gain_log, y=income)))+geom_point()+geom_smooth(method="loess",span=1)+
  labs(title = "Income by Log Capital Gain", x = "Capital Gain", y = "Income")
ggplot(adult_dum,(aes(x=capital_loss_log, y=income)))+geom_point()+geom_smooth(method="loess",span=1)+
  labs(title = "Income by Log Capital Loss", x = "Capital Loss", y = "Income")
```






```{r eda_gender}
ggplot(adult, aes(x = gender, y = age, colour = income, fill = income)) +
  geom_boxplot(color='black') +
  ggtitle("Earnings by Age and Gender")

ggplot(adult, aes(x = gender, y = years_education, colour = income, fill = income)) +
  geom_boxplot(color='black') +
  ggtitle("Earnings by Gender and Education")

ggplot(adult, aes(x = gender, y = hours_per_week, colour = income, fill = income)) +
  geom_boxplot(color='black') +
  ggtitle("Earnings by Hours Worked and Gender")

ggplot(adult, aes(x = income, y = hours_per_week, colour = income, fill = income)) +
  geom_boxplot(color='black') +
  ggtitle("Earnings by Hours Worked")

# Income by gender and hourswith percentage
ggplot(adult, aes(x = age, fill = income)) +
  geom_bar(width = 1, position = "fill") +
  facet_grid(. ~ gender) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(y = "Percentage")+
  ggtitle("Earnings by Age Type and Gender") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))

# Income by gender and hourswith percentage
ggplot(adult, aes(x = hours_per_week, fill = income)) +
  geom_bar(width = 1, position = "fill") +
  facet_grid(. ~ gender) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(y = "Percentage")+
  ggtitle("Earnings by Hours Worked Type and Gender") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))

# Income by gender and education with percentage
ggplot(adult, aes(x = years_education, fill = income)) +
  geom_bar(width = 0.8, position = "fill") +
  facet_grid(. ~ gender) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(y = "Percentage")+
  ggtitle("Earnings by Years Education and Gender") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))


```


```{r loess_gender}
# Check the continuous variables for linearity against the logodds of Y
levels(adult$gender)
# Adjusted code for Age (AG) by weight (WT), with y = Status.Num
ggplot(adult_dum, aes(x=age, y=income, colour=as.factor(gender_num))) + 
  geom_point() +
  geom_smooth(method="loess", size=1, span=.75) +
  ylim(-.2, 1.2) +
  # facet_wrap(~factor(Pclass)) +
  labs(title = "Income by Age with Gender", x = "Age", y = "Income") +
      scale_colour_manual(name = "Gender",
                        values = c("1" = "orange",  
                                   "2" = "green"),
                      labels = c("Female", "Male")) 

ggplot(adult_dum, aes(x=hours_per_week, y=income, colour=as.factor(gender_num))) + 
  geom_point() +
  geom_smooth(method="loess", size=1, span=.75) +
  ylim(-.2, 1.2) +
  # facet_wrap(~factor(Pclass)) +
  labs(title = "Income by hours Worked with Gender", x = "Hours", y = "Income") +
      scale_colour_manual(name = "Gender",
                        values = c("1" = "orange",  
                                   "2" = "green"),
                      labels = c("Female", "Male")) 

ggplot(adult_dum, aes(x=years_education, y=income, colour=as.factor(gender_num))) + 
  geom_point() +
  geom_smooth(method="loess", size=1, span=.75) +
  ylim(-.2, 1.2) +
  # facet_wrap(~factor(Pclass)) +
  labs(title = "Income by Years of Education with Gender", x = "Years", y = "Income")+
      scale_colour_manual(name = "Gender",
                        values = c("1" = "orange",  
                                   "2" = "green"),
                      labels = c("Female", "Male")) 


```




```{r eda_race}
ggplot(adult, aes(x = race, y = age, colour = income, fill = income)) +
  geom_boxplot(color='black') +
  ggtitle("Earnings by Age and Race")

ggplot(adult, aes(x = race, y = years_education, colour = income, fill = income)) +
  geom_boxplot(color='black') +
  ggtitle("Earnings by Education and Race")

ggplot(adult, aes(x = race, y = hours_per_week, colour = income, fill = income)) +
  geom_boxplot(color='black') +
  ggtitle("Earnings by Hours Worked and Race")

# Income by occupation
qplot(occupation, data = adult, fill = income) + facet_grid (. ~ race)+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))

# Income by race with percentage
ggplot(adult, aes(x = occupation, fill = income)) +
  geom_bar(position = "fill") +
  facet_grid(. ~ race) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(y = "Percentage")+
  ggtitle("Earnings by Occupation and Race") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))

# Income by race
qplot(education_level, data = adult, fill = income) +
  facet_grid(. ~ race) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))

# Income by race with percentage
ggplot(adult, aes(x = education_level, fill = income)) +
  geom_bar(position = "fill") +
  facet_grid(. ~ race) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(y = "Percentage")+
  ggtitle("Earnings by Education Type and Race") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))

# Income by race with percentage
ggplot(adult, aes(x = marital_status, fill = income)) +
  geom_bar(position = "fill") +
  facet_grid(. ~ race) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(y = "Percentage")+
  ggtitle("Earnings by Marital Status and Race (Percentage") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))

# Hourse worked
ggplot(adult, aes(x = occupation, y = hours_per_week, colour = income, fill = income)) +
  geom_boxplot(color='black') +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))+
  facet_grid(. ~ race)+
  ggtitle("Earnings by Race, Occupation, and Hours Worked")


```


```{r loess_race}
# Check the continuous variables for linearity against the logodds of Y

# Adjusted code for Age (AG) by weight (WT), with y = Status.Num
ggplot(adult_dum, aes(x=age, y=income, colour=as.factor(race_num))) + 
  geom_point() +
  geom_smooth(method="loess", size=1, span=.75) +
  ylim(-.2, 1.2) +
  # facet_wrap(~factor(Pclass)) +
  labs(title = "Income by Age with Race", x = "Age", y = "Income") +
      scale_colour_manual(name = "Race",
                        values = c("1" = "orange",  
                                   "2" = "green",  
                                   "3" = "#619CFF",  
                                   "4" = "#F564E3",  
                                   "5" = "red"),
                      labels = c("Other", "White", "Black", "Native", "Asian/Pacific Islander")) 

ggplot(adult_dum, aes(x=hours_per_week, y=income, colour=as.factor(race_num))) + 
  geom_point() +
  geom_smooth(method="loess", size=1, span=.75) +
  ylim(-.2, 1.2) +
  # facet_wrap(~factor(Pclass)) +
  labs(title = "Income by hours Worked with Race", x = "Hours", y = "Income") +
    scale_colour_manual(name = "Race",
                        values = c("1" = "orange",  
                                   "2" = "green",  
                                   "3" = "#619CFF",  
                                   "4" = "#F564E3",  
                                   "5" = "red"),
                      labels = c("Other", "White", "Black", "Native", "Asian/Pacific Islander")) 

ggplot(adult_dum, aes(x=years_education, y=income, colour=as.factor(race_num))) + 
  geom_point() +
  geom_smooth(method="loess", size=1, span=.75) +
  ylim(-.2, 1.2) +
  # facet_wrap(~factor(Pclass)) +
  labs(title = "Income by Years of Education with Race", x = "Years", y = "Income")+
    scale_colour_manual(name = "Race",
                        values = c("1" = "orange",  
                                   "2" = "green",  
                                   "3" = "#619CFF",  
                                   "4" = "#F564E3",  
                                   "5" = "red"),
                      labels = c("Other", "White", "Black", "Native", "Asian/Pacific Islander")) 


```




```{r eda_marital_status}

# Income by gender
qplot(race, data = adult, fill = income) +
  facet_grid(. ~ marital_status) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))+
  ggtitle('Earnings with Race and Marital Status')

# Income by gender
qplot(education_level, data = adult, fill = income) +
  facet_grid(. ~ marital_status) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))+
  ggtitle('Earnings with Education and Marital Status')

# Income by gender
qplot(occupation, data = adult, fill = income) +
  facet_grid(. ~ marital_status) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))+
  ggtitle('Earnings with Occupation and Marital Status')

```


```{r loess_marital_status}
# Check the continuous variables for linearity against the logodds of Y
levels(adult$marital_status)
# Adjusted code for Age (AG) by weight (WT), with y = Status.Num
ggplot(adult_dum, aes(x=age, y=income, colour=as.factor(marital_status_num))) + 
  geom_point() +
  geom_smooth(method="loess", size=1, span=.75) +
  ylim(-.2, 1.2) +
  # facet_wrap(~factor(Pclass)) +
  labs(title = "Income by Age with Marital Status", x = "Age", y = "Income") +
      scale_colour_manual(name = "Marital Status",
                        values = c("1" = "orange",  
                                   "2" = "green",  
                                   "3" = "#619CFF",  
                                   "4" = "#F564E3"),
                      labels = c("divorced", "married", "notmarried", "widowed")) 

ggplot(adult_dum, aes(x=hours_per_week, y=income, colour=as.factor(marital_status_num))) + 
  geom_point() +
  geom_smooth(method="loess", size=1, span=.75) +
  ylim(-.2, 1.2) +
  # facet_wrap(~factor(Pclass)) +
  labs(title = "Income by hours Worked with Marital Status", x = "Hours", y = "Income") +
      scale_colour_manual(name = "Marital Status",
                        values = c("1" = "orange",  
                                   "2" = "green",  
                                   "3" = "#619CFF",  
                                   "4" = "#F564E3"),
                      labels = c("divorced", "married", "notmarried", "widowed")) 

ggplot(adult_dum, aes(x=years_education, y=income, colour=as.factor(marital_status_num))) + 
  geom_point() +
  geom_smooth(method="loess", size=1, span=.75) +
  ylim(-.2, 1.2) +
  # facet_wrap(~factor(Pclass)) +
  labs(title = "Income by Years of Education with Marital Status", x = "Years", y = "Income")+
      scale_colour_manual(name = "Marital Status",
                        values = c("1" = "orange",  
                                   "2" = "green",  
                                   "3" = "#619CFF",  
                                   "4" = "#F564E3"),
                      labels = c("divorced", "married", "notmarried", "widowed")) 


```




```{r eda_mixed_variables}
ggplot(adult, aes(x = hours_worked_span, y = age, colour = income, fill = income)) +
  geom_boxplot(color='black') +
  ggtitle("Earnings by Hours Worked")



ggplot(adult, aes(x = income, y = age, color=income, fill=income)) + 
  geom_boxplot(color='black') +
  facet_grid(. ~ native_country) +
  ggtitle("Earnings by Age and Origin")

ggplot(adult) + aes(x=hours_per_week, group=income, fill=income) + 
  geom_histogram()+
  facet_grid(. ~ occupation) 

# Income by race with percentage
ggplot(adult, aes(x = hours_worked_span, fill = income)) +
  geom_bar(position = "fill") +
  facet_grid(. ~ occupation) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(y = "Percentage")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))

# Income by race with percentage
ggplot(adult, aes(x = native_country, fill = income)) +
  geom_bar(position = "fill") +
  facet_grid(. ~ occupation) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(y = "Percentage")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))


ggplot(adult, aes(x = marital_status, y = hours_per_week, colour = income, fill = income)) +
  geom_boxplot(color='black') +
  ggtitle("Earnings by Marital Status and Education")
```




```{r eda_more_mixed_variables}
# Hourse worked
ggplot(adult, aes(x = gender, y = hours_per_week, colour = income, fill = income)) +
  geom_boxplot(color='black') +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))+
  facet_grid(. ~ race)+
  ggtitle("Earnings by Race, Gender, and Hours Worked")

ggplot(adult, aes(x = gender, y = years_education, colour = income, fill = income)) +
  geom_boxplot(color='black') +
  ggtitle("Earnings by Gender and Education")

ggplot(adult, aes(x = marital_status, y = hours_per_week, colour = income, fill = income)) +
  geom_boxplot(color='black') +
  ggtitle("Earnings by Marital Status and Hours")

# Income by gender
qplot(race, data = adult, fill = income) +
  facet_grid(. ~ marital_status) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))+
  ggtitle('Earnings with Race and Marital Status')


```



```{r eda_occupation_education}
# Income by occupation
qplot(income, data = adult, fill = income) + facet_grid (. ~ occupation)+
  ggtitle('Earnings with Occupation')

# Income by education
qplot(income, data = adult, fill = income) + facet_grid (. ~ education_level)+
  ggtitle('Earnings with Education')

# Income by race
qplot(education_level, data = adult, fill = income) +
  facet_grid(. ~ race) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))+
  ggtitle('Earnings with Race and Education')


```


```{r eda_native_region}
# Load the necessary package (if not already installed)
library(vcd)

# Create a mosaic plot for the "region" variable
mosaic(~ native_country, data = adult, shade = TRUE, legend = TRUE)


library(treemap)

levels(adult_dum$income)
adult_dum$income<-as.numeric(adult_dum$income)
positive_proportions <- aggregate(income ~ native_country_num, data = adult_dum, FUN = mean)

treemap(positive_proportions, index = "native_country_num", vSize = "income", title = "Proportion of Positive Income Status by native_country of Origin")



# Create a treemap plot with labels
library(treemap)

# Define the labels for each native country
labels <- c("EAsia", "SAsia", "North America", "Central America", "South America", "Europe", "US Territory")

# Aggregate income proportions by native_country_num
positive_proportions <- aggregate(income ~ native_country_num, data = adult_dum, FUN = mean)

# Create the treemap
treemap(positive_proportions, index = "native_country_num", vSize = "income",
        title = "Proportion of Positive Income Status by Native Country of Origin",
        draw = FALSE) #+
    # geom_treemap() +  # Add the treemap
    # geom_text(aes(label = labels), color = "white")  # Add labels




# Heat map
income_status_table <- table(adult$native_country, adult_dum$income)

positive_counts <- income_status_table[, "1"]
total_counts <- rowSums(income_status_table)

positive_proportions<-positive_counts/total_counts

heatmap_data <- data.frame(native_country = names(positive_proportions), positive_proportion = positive_proportions)

heatmap_data$native_country <- factor(heatmap_data$native_country, levels = heatmap_data$native_country[order(heatmap_data$positive_proportion, decreasing = FALSE)])


ggplot(heatmap_data, aes(x = "", y = native_country, fill = positive_proportion)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "blue") +
  theme_minimal() +
  labs(title = "Proportion of Positive Income Status by native_country of Origin", x = NULL, y = NULL) +
        scale_colour_manual(name = "Region",
                        values = c("1" = "orange",  
                                   "2" = "green",  
                                   "3" = "#619CFF",  
                                   "4" = "#F564E3",  
                                   "5" = "red",
                                   "6" = "purple",
                                   "7" = "brown"),
                      labels = c("EAsia", "SAsia", "North America", "Central America", "South America", "Europe", "US Territory")) 
```





```{r loess_region, WARNING = FALSE}
# Check the continuous variables for linearity against the logodds of Y

# Adjusted code for Age (AG) by weight (WT), with y = Status.Num
ggplot(adult_dum, aes(x=capital_gain_log, y=income, colour=as.factor(native_country_num))) + 
  geom_point() +
  geom_smooth(method="loess", size=1, span=2) +
  ylim(-2, 1.2) +
  # facet_wrap(~factor(Pclass)) +
  labs(title = "Income by Capital Gain with Region", x = "Gain", y = "Income") +
      scale_colour_manual(name = "Region",
                        values = c("1" = "orange",  
                                   "2" = "green",  
                                   "3" = "#619CFF",  
                                   "4" = "#F564E3",  
                                   "5" = "red",
                                   "6" = "purple",
                                   "7" = "brown"),
                      labels = c("EAsia", "SAsia", "North America", "Central America", "South America", "Europe", "US Territory")) 

# Adjusted code for Age (AG) by weight (WT), with y = Status.Num
ggplot(adult_dum, aes(x=capital_loss_log, y=income, colour=as.factor(native_country_num))) + 
  geom_point() +
  geom_smooth(method="loess", size=1, span=2) +
  ylim(-2, 1.2) +
  # facet_wrap(~factor(Pclass)) +
  labs(title = "Income by Capital Loss with Region", x = "Loss", y = "Income") +
    scale_colour_manual(name = "Region",
                        values = c("1" = "orange",  
                                   "2" = "green",  
                                   "3" = "#619CFF",  
                                   "4" = "#F564E3",  
                                   "5" = "red",
                                   "6" = "purple",
                                   "7" = "brown"),
                      labels = c("EAsia", "SAsia", "North America", "Central America", "South America", "Europe", "US Territory")) 


# Adjusted code for Age (AG) by weight (WT), with y = Status.Num
ggplot(adult_dum, aes(x=years_education, y=income, colour=as.factor(native_country_num))) + 
  geom_point() +
  geom_smooth(method="loess", size=1, span=2) +
  ylim(-2, 1.2) +
  # facet_wrap(~factor(Pclass)) +
  labs(title = "Income by Years of Education with Region", x = "Years", y = "Income") +
    scale_colour_manual(name = "Region",
                        values = c("1" = "orange",  
                                   "2" = "green",  
                                   "3" = "#619CFF",  
                                   "4" = "#F564E3",  
                                   "5" = "red",
                                   "6" = "purple",
                                   "7" = "brown"),
                      labels = c("EAsia", "SAsia", "North America", "Central America", "South America", "Europe", "US Territory")) 


# Adjusted code for Age (AG) by weight (WT), with y = Status.Num
ggplot(adult_dum, aes(x=occupation_num, y=income, colour=as.factor(native_country_num))) + 
  geom_point() +
  geom_smooth(method="loess", size=1, span=2) +
  ylim(-2, 1.2) +
  # facet_wrap(~factor(Pclass)) +
  labs(title = "Income by Occupation with Region", x = "Occupation", y = "Income") +
    scale_colour_manual(name = "Region",
                        values = c("1" = "orange",  
                                   "2" = "green",  
                                   "3" = "#619CFF",  
                                   "4" = "#F564E3",  
                                   "5" = "red",
                                   "6" = "purple",
                                   "7" = "brown"),
                      labels = c("EAsia", "SAsia", "North America", "Central America", "South America", "Europe", "US Territory")) 



ggplot(adult_dum,(aes(x=capital_gain_log,y=income, colour=as.factor(race_num))))+geom_point()+geom_smooth(method="loess",span=2) +
    ylim(-2, 1.2) +
  # facet_wrap(~factor(Pclass)) +
  labs(title = "Income by capital gain with Race", x = "Gain", y = "Income") +
  scale_colour_manual(name = "Race",
                        values = c("1" = "orange",  
                                   "2" = "green",  
                                   "3" = "#619CFF",  
                                   "4" = "#F564E3",  
                                   "5" = "red"),
                      labels = c("Other", "White", "Black", "Native", "Asian/Pacific Islander"))   




```



```{r drop_unused_variable}
adult2 <- adult
# Log-transform the 'capital_gain' column
adult2$capital_gain_log <- log(adult2$capital_gain + 1)  # Adding 1 to avoid log(0)

# Log-transform the 'capital_gain' column
adult2$capital_loss_log <- log(adult2$capital_loss + 1)  # Adding 1 to avoid log(0)

# Drop the specified columns
adult2 <- subset(adult2, select = -c(age_span, hours_worked_span, similar_pop_count, capital_gain, capital_loss, invested))

```




#Check Logistic Assumptions:
Sources: https://www.statology.org/assumptions-of-logistic-regression/
* https://www.statisticssolutions.com/free-resources/directory-of-statistical-analyses/assumptions-of-logistic-regression/
* https://r4ds.github.io/bookclub-islr/addendum---logistic-regression-assumptions.html
* https://sscc.wisc.edu/sscc/pubs/RegDiag-R/logistic-regression.html#log_ind
* http://www.sthda.com/english/articles/36-classification-methods-essentials/148-logistic-regression-assumptions-and-diagnostics-in-r/#google_vignette
* https://towardsdatascience.com/assumptions-of-logistic-regression-clearly-explained-44d85a22b290
* Assumption 1: Is the response variable binary?
```{r binary_check}
# Assure it is a factor
adult2$income <- factor(adult2$income, levels = c("<=50K", ">50K"))
levels(adult2$income)

# Visual inspection (example with 'age' as a predictor)
ggplot(adult2, aes(x = age, fill = income)) + geom_histogram(position = "dodge")

```


# Assumption 2: Independence of Observations
* The Durbin-Watson statistic (DW) is close to 2 (DW = 1.9966), which indicates no autocorrelation, and the high p-value (p-value = 0.3786) means that you fail to reject the null hypothesis of no autocorrelation
```{r independence_check}
# Fit the logistic regression model
model <- glm(income ~ ., data = adult2, family = binomial)

# Perform the Durbin-Watson test
dw_result <- dwtest(model)
dw_result

```

# Assumption 3: No Multicollinearity among predictors.
* All Variance Inflation Factors (VIFs) are below 5 and visually verified as not being highly correlated.
* Note: a VIF value below 5 suggests that there is no significant multicollinearity among the predictors, and values below 10 are often considered acceptable.
```{r multicollinearity_check}
adult_dum <- adult2
# one-hot-encode income
adult_dum$income <- ifelse(adult_dum$income == "<=50K", 0, 1)

# Loop through columns
for (col in names(adult_dum)) {
  if (is.factor(adult_dum[[col]])) { # Check if column is a factor
    levels <- levels(adult_dum[[col]]) # Get levels of the factor
    adult_dum[[col]] <- as.numeric(factor(adult_dum[[col]], levels = levels)) # Convert factor to numeric based on levels
    names(adult_dum)[names(adult_dum) == col] <- paste0(col, "_num") # Rename the column
  }
}

adult_dum


# Fit logistic regression model
model2 <- glm(income ~ ., family = binomial, data = adult_dum)

# Calculate VIF
vif_values <- vif(model2)
vif_values

# Visualize correlation matrix of predictors
# Visualize correlation matrix of predictors
cor_matrix <- cor(dplyr::select(adult_dum, -income))
corrplot(cor_matrix, method='circle')

```


# Assumption 4: Check for Extreme Outliers with high influence
* There are 28 outliers that need to be removed.
* Note: Data points with an absolute standardized residuals above 3 represent possible outliers and may deserve closer attention.
```{r}
library(broom)
model <- glm(income ~ ., data = adult2, family = binomial)
plot(model, which = 4, id.n = 10)

# Extract model results
model.data <- augment(model) %>% 
  dplyr::mutate(index = dplyr::row_number())  # Use row_number() instead of n()

# Show the top 5
model.data %>% top_n(10, .cooksd)

# Cook's D influencial outliers
ggplot(model.data, aes(index, .std.resid)) + 
  geom_point(aes(color = income), alpha = 0.5) +
  theme_bw()

# Influence Plot and list of top influencers - Continuous
influencePlot(model, id=list(method="noteworthy", n=5, cex=.75, col=carPalette()[1], location="lr"), 
              main="Influence Plot", sub="Circle area is proportial to Cook's Distance - Continuous" )

# Remove influential data points
model.data %>% 
  filter(abs(.std.resid) > 3)

# Remove influential data points based on standardized residuals
adult.out <- adult2 %>% 
  slice(-which(abs(model.data$.std.resid) > 3))

model_out <- glm(income ~ ., data = adult.out, family = binomial)
plot(model_out, which = 4, id.n = 10)

# Influence Plot and list of top influencers - Continuous
influencePlot(model_out, id=list(method="noteworthy", n=5, cex=.75, col=carPalette()[1], location="lr"), 
              main="Influence Plot", sub="Circle area is proportial to Cook's Distance - Continuous" )
```

# Assumption 5: There is a Linear Relationship Between Explanatory (continuous) Variables and the Logit (log odds) of the Response Variable 
* For categorical predictors, the concept of linearity doesn’t apply in the traditional sense because these variables are represented in the model as a series of dummy variables (one for each category, minus one reference category).
* Because the dependent variable is categorical and any transformation to dummy variables would not be an accurate assessment of a linear relationship, you don't have to do this assumptions check on the categorical. Below are just some tests.
* Sources: https://bookdown.org/sarahwerth2024/CategoricalBook/logistic-regression-r.html
```{r}
library(tidyverse)
library(broom)


# Fit the logistic regression model
model <- glm(income ~ ., data = adult_dum, family = binomial)

# Predict the probability (p) of diabetes positivity
probabilities <- predict(model, type = "response")

# Create binary predicted classes based on a threshold (e.g., 0.5)
predicted_classes <- ifelse(probabilities > 0.24, "0", "1")

# Visualize the relationship between a continuous predictor (e.g., age) and logit values
logit_values <- log(probabilities / (1 - probabilities))

# All One-Hot-Encoded variables
ggplot(adult_dum,(aes(x=logit_values,y=income)))+geom_point()+geom_smooth(method="loess",span=.9)

# logit against age
ggplot(adult_dum,(aes(x=logit_values,y=age)))+geom_point()+geom_smooth(method="loess",span=.9)

# logit against years_education
ggplot(adult_dum,(aes(x=logit_values,y=years_education)))+geom_point()+geom_smooth(method="loess",span=.9)

# Check for linearity in continuous variables
boxTidwell(income ~ age + years_education, data = adult_dum)

```




# Post EDA/assumption check data-split to train and test sets
```{r create_train_test}
# Set a seed for reproducibility
set.seed(1234)

# Calculate the number of rows to select
sample_size <- round(0.8 * nrow(adult.out))

# Create a random sample of indices
index <- sample(1:nrow(adult2), size = sample_size, replace = FALSE)

# Create the training set using the selected indices
training_set <- adult2[index, ]

# Create the testing set with the remaining indices
testing_set <- adult2[-index, ]
nrow(training_set)
nrow(testing_set)

testing_set

# For continuous check
index <- sample(1:nrow(adult_dum), size = round(nrow(adult_dum) * 0.8))

# Splitting
training_set_LR <- adult_dum[index, ]
testing_set_LR <- adult_dum[-index, ]

#Outliers removed
index <- sample(1:nrow(adult.out), size = round(nrow(adult.out) * 0.8))

# Splitting
training_set_out <- adult.out[index, ]
testing_set_out <- adult.out[-index, ]
```

# Initial test of the model without changes
* Result summary: A review of the 1) full logistic, 2) influential outliers removed, and 3) the One-Hot-Encoded variables models show slight difference in the confusion matrix results and the significance levels. Model4 outperforms model3 and model5 slightly, and all tests have similar AUROC scores and achieve a balance of approximately 80 percent on Sensitivity and Specificity. The one-hot-encoded model found many more variables to be significant, but had lower performance overall.
```{r initial_model, warning=FALSE}

# Fit a logistic regression model
model3 <- glm(income ~ ., data = training_set, family = binomial)

# Confidence interval
# confint(model3)

# Summarize the final selected model
summary(model3)

# Check for multicollinearity
vif(model3) 

# Perform Hosmer-Lemeshow goodness of fit test
# Note: A p-value less than 0.05 typically suggests that the model does not fit the data well.
hoslem.test(training_set$income, fitted(model3))
# hoslem.test(training_set_LR$income, fitted(model4)) # dummied variables - test comparison

# Make predictions
probabilities <- model3 %>% predict(testing_set, type = "response")
predicted.classes <- ifelse(probabilities > 0.26, ">50K", "<=50K")

# Model accuracy
accuracy <- mean(predicted.classes==testing_set$income)

# confusion matrix 
tb <- table(predicted.classes, testing_set$income)
tb

# Calculate Sensitivity (True Positive Rate)
sensitivity <- tb[2, 2] / (tb[2, 2] + tb[1, 2])

# Calculate Specificity (True Negative Rate)
specificity <- tb[1, 1] / (tb[1, 1] + tb[2, 1])

# Calculate Prevalence
prevalence <- (tb[1, 2] + tb[2, 2]) / sum(tb)

# Calculate PPV (Positive Predictive Value)
ppv <- tb[2, 2] / (tb[2, 2] + tb[2, 1])

# Calculate NPV (Negative Predictive Value)
npv <- tb[1, 1] / (tb[1, 1] + tb[1, 2])

# Calculate AUROC
roc_obj <- roc(testing_set$income, probabilities)
auroc <- auc(roc_obj)

# Print the metrics
cat("accuracy:", accuracy, "\n")
cat("Sensitivity:", sensitivity, "\n")
cat("Specificity:", specificity, "\n")
cat("Prevalence:", prevalence, "\n")
cat("PPV:", ppv, "\n")
cat("NPV:", npv, "\n")
cat("AUROC:", auroc, "\n")



################## glm test with influential outliers removed ##############################
cat("\n\n----------------influential outliers removed--------------\n\n")

# Fit a logistic regression model
model4 <- glm(income ~ ., data = training_set_out, family = binomial)

# Confidence interval
# confint(model3)

# Summarize the final selected model
summary(model4)

# Check for multicollinearity
vif(model4) 

# Perform Hosmer-Lemeshow goodness of fit test
# Note: A p-value less than 0.05 typically suggests that the model does not fit the data well.
hoslem.test(training_set_out$income, fitted(model4))
# hoslem.test(training_set_LR$income, fitted(model4)) # dummied variables - test comparison

# Make predictions
probabilities2 <- model4 %>% predict(testing_set_out, type = "response")
predicted.classes2 <- ifelse(probabilities2 > 0.275, ">50K", "<=50K")

# Model accuracy
accuracy2 <- mean(predicted.classes2==testing_set_out$income)

# confusion matrix 
tb2 <- table(predicted.classes2, testing_set_out$income)
tb2

# Calculate Sensitivity (True Positive Rate)
sensitivity2 <- tb2[2, 2] / (tb2[2, 2] + tb2[1, 2])

# Calculate Specificity (True Negative Rate)
specificity2 <- tb2[1, 1] / (tb2[1, 1] + tb2[2, 1])

# Calculate Prevalence
prevalence2 <- (tb2[1, 2] + tb2[2, 2]) / sum(tb2)

# Calculate PPV (Positive Predictive Value)
ppv2 <- tb2[2, 2] / (tb2[2, 2] + tb2[2, 1])

# Calculate NPV (Negative Predictive Value)
npv2 <- tb2[1, 1] / (tb2[1, 1] + tb2[1, 2])

# Calculate AUROC
roc_obj2 <- roc(testing_set_out$income, probabilities2)
auroc2 <- auc(roc_obj2)

# Print the metrics
cat("accuracy:", accuracy2, "\n")
cat("Sensitivity:", sensitivity2, "\n")
cat("Specificity:", specificity2, "\n")
cat("Prevalence:", prevalence2, "\n")
cat("PPV:", ppv2, "\n")
cat("NPV:", npv2, "\n")
cat("AUROC:", auroc2, "\n")


# full with outliers
pred_full <- prediction(probabilities, testing_set$income)
pred_full_per <- performance(pred_full, measure = "tpr", x.measure = "fpr")
mod1_pred <- data.frame(FP = pred_full_per@x.values[[1]], TP = pred_full_per@y.values[[1]])

# full without outliers
pred_full_Out <- prediction(probabilities2, testing_set_out$income)
pred_full_per2 <- performance(pred_full_Out, measure = "tpr", x.measure = "fpr")
mod2_pred <- data.frame(FP = pred_full_per2@x.values[[1]], TP = pred_full_per2@y.values[[1]])

# plot ROC curve for logistic regression
g <- ggplot() + 
  geom_line(data = mod1_pred, aes(x = FP, y = TP, color = 'Full With Outliers')) + 
  geom_line(data = mod2_pred, aes(x = FP, y = TP, color = 'Full Without Outliers')) +
  geom_segment(aes(x = 0, xend = 1, y = 0, yend = 1)) +
  ggtitle('ROC Curve') + 
  labs(x = 'False Positive Rate', y = 'True Positive Rate') 


g +  scale_colour_manual(name = 'Classifier', values = c('Full With Outliers'='#E69F00', 
                                               'Full Without Outliers'='#56B4E9'))


######## Base test with all Categorical variables changed to Dummies/Continuous ##########
cat("\n\n----------------Variables changed to Continuous Dummies--------------\n\n")

# Fit a logistic regression model
model5 <- glm(income ~ ., data = training_set_LR, family = binomial)

# Confidence interval
# confint(model3)

# Summarize the final selected model
summary(model5)

# Check for multicollinearity
vif(model5) 

# Perform Hosmer-Lemeshow goodness of fit test
# Note: A p-value less than 0.05 typically suggests that the model does not fit the data well.
#hoslem.test(training_set$income, fitted(model5))
# hoslem.test(training_set_LR$income, fitted(model4)) # dummied variables - test comparison

# Make predictions
probabilities <- model5 %>% predict(testing_set_LR, type = "response")
predicted.classes2 <- ifelse(probabilities > 0.27, "1", "0")

# Model accuracy
accuracy <- mean(predicted.classes2==testing_set_LR$income)

# confusion matrix 
tb <- table(predicted.classes2, testing_set_LR$income)
tb

# Calculate Sensitivity (True Positive Rate)
sensitivity <- tb[2, 2] / (tb[2, 2] + tb[1, 2])

# Calculate Specificity (True Negative Rate)
specificity <- tb[1, 1] / (tb[1, 1] + tb[2, 1])

# Calculate Prevalence
prevalence <- (tb[1, 2] + tb[2, 2]) / sum(tb)

# Calculate PPV (Positive Predictive Value)
ppv <- tb[2, 2] / (tb[2, 2] + tb[2, 1])

# Calculate NPV (Negative Predictive Value)
npv <- tb[1, 1] / (tb[1, 1] + tb[1, 2])

# Calculate AUROC
roc_obj <- roc(testing_set_LR$income, probabilities)
auroc <- auc(roc_obj)

# Print the metrics
cat("accuracy:", accuracy, "\n")
cat("Sensitivity:", sensitivity, "\n")
cat("Specificity:", specificity, "\n")
cat("Prevalence:", prevalence, "\n")
cat("PPV:", ppv, "\n")
cat("NPV:", npv, "\n")
cat("AUROC:", auroc, "\n")

```

# Odds ratio for the model removing influencial outliers.
An OR of x suggests that with each additional year of (the variable), the odds of the outcome occurring increase by approximately ((OR -1)*100) all other variables remaining constant. or if it has a reference variable...compared to the reference variable.
```{r warning=FALSE}
summary(model4)
# Calculate Odds Ratios and 90% Confidence Intervals
odds_ratios <- exp(coef(model4))
conf_int <- exp(confint(model4, level = 0.95))

# Identify significant predictors at the 0.05 level
significant_predictors <- summary(model4)$coefficients[, 4] < 0.05

# Extract Odds Ratios and Confidence Intervals for significant predictors
# Note: Use names() to match significant predictors by name
significant_or_ci <- data.frame(
  OddsRatio = odds_ratios[significant_predictors],
  LowerCI = conf_int[names(odds_ratios)[significant_predictors], 1],
  UpperCI = conf_int[names(odds_ratios)[significant_predictors], 2]
)
significant_or_ci

```

# Interpretation:

age: An odds ratio (OR) of 1.02560964 means that with each additional year of age, the odds of the outcome occurring increase by approximately 2.56%. The calculation is ((OR - 1) \times 100 = (1.02560964 - 1) \times 100 = 2.56%).
similar_pop_count: An OR of 1.00000063 indicates an extremely small increase in odds with each unit increase in similar population count. The effect is so small it’s practically negligible.
years_education: An OR of 1.27846414 suggests that with each additional year of education, the odds of the outcome occurring increase by 27.85%. The calculation is ((1.27846414 - 1) \times 100 = 27.85%).
marital_statusmarried: An OR of 1.83198084 means that being married is associated with an 83.20% increase in the odds of the outcome occurring compared to not being married.
marital_statusnotmarried: An OR of 0.55312175 indicates that not being married is associated with a 44.69% decrease in the odds of the outcome occurring compared to being married. The calculation is ((1 - 0.55312175) \times 100 = 44.69%).
occupationclerical: An OR of 2.73417509 suggests that being in a clerical occupation is associated with a 173.42% increase in the odds of the outcome occurring compared to not being in this occupation.
occupationhighskill: An OR of 4.87181987 indicates that being in a high-skill occupation is associated with a 387.18% increase in the odds of the outcome occurring compared to not being in this occupation.
occupationmidskill: An OR of 2.61052015 means that being in a mid-skill occupation is associated with a 161.05% increase in the odds of the outcome occurring compared to not being in this occupation.
household_rolewife: An OR of 4.11794882 suggests that being the wife in a household is associated with a 311.79% increase in the odds of the outcome occurring compared to not being the wife.
household_roletenant: An OR of 0.34585959 indicates that being a tenant is associated with a 65.41% decrease in the odds of the outcome occurring compared to not being a tenant.
household_rolesingle_occ: An OR of 0.29151166 means that being a single occupant is associated with a 70.85% decrease in the odds of the outcome occurring compared to not being a single occupant.
household_rolerelative: An OR of 0.23532074 suggests that being a relative in a household is associated with a 76.47% decrease in the odds of the outcome occurring compared to not being a relative.
household_roledepchild: An OR of 0.11687608 indicates that being a dependent child is associated with an 88.31% decrease in the odds of the outcome occurring compared to not being a dependent child.
gendermale: An OR of 2.45071926 means that being male is associated with a 145.07% increase in the odds of the outcome occurring compared to not being male.
capital_gain: An OR of 1.00097874 suggests a very slight increase in the odds of the outcome occurring with each unit increase in capital gain.
capital_loss: An OR of 1.00246863 indicates a slight increase in the odds of the outcome occurring with each unit increase in capital loss.
hours_per_week: An OR of 1.02929197 means that with each additional hour worked per week, the odds of the outcome occurring increase by approximately 2.93%.
native_countrySouthAmerica: An OR of 0.19791245 suggests that being from South America is associated with an 80.21% decrease in the odds of the outcome occurring compared to not being from South America.
investedInvested: An OR of 0.02628912 indicates that having invested is associated with a 97.37% decrease in the odds of the outcome occurring compared to not having invested.






Drop for code or run testing_set as well:

# Build a reduced model from the significant variables
* Test cost for automating extracting the reduced model.
* The testing set would also need to be refined for this formula.
```{r reduced_model}
# Get the original variable names (excluding the response variable)
original_vars <- setdiff(names(training_set), "income")

# Summary of the full model
summary_full <- summary(model3)

# Extracting significant variables
significant_vars <- summary_full$coefficients[summary_full$coefficients[, "Pr(>|z|)"] < 0.05, ]
significant_vars_names <- rownames(significant_vars)

# Initialize an empty string to build the new formula
new_formula <- "income ~ "

# Create a list to keep track of which categorical variables have all non-reference levels significant
all_levels_significant <- list()

# Loop through each original variable to check if all non-reference levels are significant
for(original_var in original_vars) {
  # Get all non-reference levels of the categorical variable
  levels <- levels(training_set[[original_var]])[-1] # Exclude the first level which is the reference
  
  # Check if all non-reference levels are significant
  all_significant <- all(sapply(levels, function(lvl) paste0(original_var, lvl) %in% significant_vars_names))
  
  # Store the result
  all_levels_significant[[original_var]] <- all_significant
}

# Loop through each significant variable
for(var in significant_vars_names) {
  # Initialize new_var_expr
  new_var_expr <- var
  
  # Check if var is a level of an original variable
  for(original_var in original_vars) {
    if(grepl(original_var, var)) {
      # Check if all non-reference levels of the variable are significant
      if(all_levels_significant[[original_var]]) {
        # If all non-reference levels are significant, use the original variable name
        new_var_expr <- original_var
      } else {
        # If not all levels are significant, construct the new variable expression
        level <- gsub(paste0("^", original_var), "", var)
        if(nchar(level) > 0) {
          new_var_expr <- paste0("(training_set$", original_var, " == '", level, "')")
        }
      }
      # Ensure only the condition is added to the formula, not the logical result
      new_formula <- paste(new_formula, new_var_expr, sep = " + ")
      break
    }
  }
}

# Remove the trailing " + " if present
new_formula <- sub(" \\+ $", "", new_formula)

# Convert the string to a formula
new_formula <- as.formula(new_formula)

# Fit the reduced model
reduced_model <- glm(new_formula, data = training_set, family = binomial)
summary(reduced_model)

# Check for multicollinearity
vif(reduced_model) 

# Perform Hosmer-Lemeshow goodness of fit test
# Note: A p-value less than 0.05 typically suggests that the model does not fit the data well.
hoslem.test(training_set$income, fitted(reduced_model))
# hoslem.test(training_set_LR$income, fitted(model4)) # dummied variables - test comparison


```


PCA Sources: https://www.analyticsvidhya.com/blog/2021/09/pca-and-its-underlying-mathematical-principles/
# PCA full model Test:
For PC1: marital_status_num and education_level have high positive loading scores, while age has a high negative loading score. These variables significantly influence PC1.
For PC2: capital_gain and capital_loss are notable due to their high positive loading scores, making them important for PC2.
For PC3: hours_per_week stands out with a high positive loading score, suggesting its relevance to PC3.
12 variables are needed for 90 percent.
```{r}
pc.result<-prcomp(training_set_LR[,1:14],scale.=TRUE)

#Eigen Vectors
pc.result$rotation

#Eigen Values
cat("\n\n----------------Eigen Values--------------\n\n")
eigenvals<-pc.result$sdev^2
eigenvals

# Calculate cumulative proportion of variance explained
cumulative_var <- cumsum(eigenvals) / sum(eigenvals)

# Find the number of PCs needed to retain 90% of the variance
num_pcs <- which(cumulative_var >= 0.9)[1]

# number of PCs
num_pcs

#Scree plot
par(mfrow=c(1,2))
plot(eigenvals,type="l",main="Scree Plot",ylab="Eigen Values",xlab="PC #")
plot(eigenvals/sum(eigenvals),type="l",main="Explained Variance",ylab = "Cumulative Proportion", xlab = "Principal Component", ylim=c(0,1))
cumulative.prop<-cumsum(eigenvals/sum(eigenvals))
lines(cumulative.prop,type = "o", lty=2)
abline(h = 0.9, col = "red", lty = 2) # Add a line at 90% variance explained
legend("topright", legend=c("Prop","Cumulative"),
       lty=1:2, cex=0.8)
```

PCA reduced Model:
PC1: household_role_num (0.49898102), invested_num (-0.38056203), age (-0.36436500)
PC2: invested_num (0.5153051), capital_loss (0.4491392), years_education (0.2559078)
PC3: occupation_num (0.63406086), capital_loss (0.36206738), invested_num (0.28760465)
```{r}

pc.result <- prcomp(training_set_LR[, c("age", "years_education", "capital_gain_log", "capital_loss_log",
                                   "marital_status_num", "occupation_num", "household_role_num", "gender_num", 
                                   "hours_per_week")], scale. = TRUE)

#Eigen Vectors
pc.result$rotation

#Eigen Values
cat("\n\n----------------Eigen Values--------------\n\n")
eigenvals<-pc.result$sdev^2
eigenvals

# Calculate cumulative proportion of variance explained
cumulative_var <- cumsum(eigenvals) / sum(eigenvals)

# Find the number of PCs needed to retain 90% of the variance
num_pcs <- which(cumulative_var >= 0.9)[1]

# number of PCs
num_pcs

#Scree plot
par(mfrow=c(1,2))
plot(eigenvals,type="l",main="Scree Plot",ylab="Eigen Values",xlab="PC #")
plot(eigenvals/sum(eigenvals),type="l",main="Explained Variance",ylab = "Cumulative Proportion", xlab = "Principal Component", ylim=c(0,1))
cumulative.prop<-cumsum(eigenvals/sum(eigenvals))
lines(cumulative.prop,type = "o", lty=2)
abline(h = 0.9, col = "red", lty = 2) # Add a line at 90% variance explained
legend("topright", legend=c("Prop","Cumulative"),
       lty=1:2, cex=0.8)
```





# GLMNET Feature Test - one variable was dropped.
```{r}
library(caret)
# change the income value names to work with glmnet (greater than sign will not run)
training_set1 <- training_set
training_set1$income <- gsub('<=50K', 'less_than_50K', training_set1$income)
training_set1$income <- gsub('>50K', 'greater_than_50K', training_set1$income)

# Convert 'income' to a factor if it's not already
training_set1$income <- as.factor(training_set1$income)
levels(training_set1$income)
# Ensure that train_control is set up for classification
train_control <- trainControl(method = "cv", number = 10, classProbs = TRUE, summaryFunction = twoClassSummary)

# Now, re-run the GLMNET model training
set.seed(1234)
glmnet_model <- train(
  income ~ ., 
  data = training_set1,  # Exclude the 'id_number' column
  method = "glmnet",
  trControl = train_control,
  preProcess = c("center", "scale"),  # Standardize predictors
  tuneLength = 10,
  metric = "ROC"  # Use AUROC curve error metric
)

# Comment on which predictors were left or dropped
coef(glmnet_model$finalModel, glmnet_model$bestTune$lambda)
```

# GLMNET Feature Test reduced_ - one variable was dropped.
```{r}
library(caret)
# change the income value names to work with glmnet (greater than sign will not run)
training_set1 <- training_set_out
training_set1$income <- gsub('<=50K', 'less_than_50K', training_set1$income)
training_set1$income <- gsub('>50K', 'greater_than_50K', training_set1$income)

# Convert 'income' to a factor if it's not already
training_set1$income <- as.factor(training_set1$income)
levels(training_set1$income)
# Ensure that train_control is set up for classification
train_control <- trainControl(method = "cv", number = 10, classProbs = TRUE, summaryFunction = twoClassSummary)

# Now, re-run the GLMNET model training
set.seed(1234)
glmnet_model <- train(
  income ~ ., 
  data = training_set1,  # Exclude the 'id_number' column
  method = "glmnet",
  trControl = train_control,
  preProcess = c("center", "scale"),  # Standardize predictors
  tuneLength = 10,
  metric = "ROC"  # Use AUROC curve error metric
)

# Comment on which predictors were left or dropped
coef(glmnet_model$finalModel, glmnet_model$bestTune$lambda)
```



# Stepwise feature selection using AIC for the step evaluation
* Sources: http://www.sthda.com/english/articles/36-classification-methods-essentials/150-stepwise-logistic-regression-essentials-in-r/
* Note: When comparing models, the one with the lower AIC is generally preferred because it suggests a better fit with fewer parameters.
* To compare models of different sizes using deviance, you need to adjust for the number of predictors, which is what AIC does by penalizing more complex models2.
```{r comparative_test, warning=FALSE}
library(MASS)
cat("\n\n--------modelFull--------\n\n") # same as model 3 (46 total variables)
full.model <- glm(income ~ ., data = training_set, family = binomial)
coef(full.model)

# Make predictions
probabilities <- full.model %>% predict(testing_set, type = "response")
predicted.classes <- ifelse(probabilities > 0.24, ">50K", "<=50K")
# Prediction accuracy
observed.classes <- testing_set$income
mean(predicted.classes == observed.classes)

summary(full.model)
# confusion matrix 
tb <- table(predicted.classes, testing_set$income)
tb

# Model accuracy
accuracy <- mean(predicted.classes==testing_set$income)

# Calculate Sensitivity (True Positive Rate)
sensitivity <- tb[2, 2] / (tb[2, 2] + tb[1, 2])

# Calculate Specificity (True Negative Rate)
specificity <- tb[1, 1] / (tb[1, 1] + tb[2, 1])

# Calculate Prevalence
prevalence <- (tb[1, 2] + tb[2, 2]) / sum(tb)

# Calculate PPV (Positive Predictive Value)
ppv <- tb[2, 2] / (tb[2, 2] + tb[2, 1])

# Calculate NPV (Negative Predictive Value)
npv <- tb[1, 1] / (tb[1, 1] + tb[1, 2])

# Calculate AUROC
roc_obj <- roc(testing_set$income, probabilities)
auroc <- auc(roc_obj)

# Print the metrics
cat("accuracy:", accuracy, "\n")
cat("Sensitivity:", sensitivity, "\n")
cat("Specificity:", specificity, "\n")
cat("Prevalence:", prevalence, "\n")
cat("PPV:", ppv, "\n")
cat("NPV:", npv, "\n")
cat("AUROC:", auroc, "\n")

cat("\n\n--------modelAICStep--------\n\n") # (35 variables recommended)
step.model <- full.model %>% stepAIC(trace = FALSE)
coef(step.model)

# Make predictions
probabilities <- predict(step.model, testing_set, type = "response")
predicted.classes <- ifelse(probabilities > 0.24, ">50K", "<=50K")

# confusion matrix 
tb <- table(predicted.classes, testing_set$income)
tb

summary(step.model)
# Model accuracy
accuracy <- mean(predicted.classes==testing_set$income)

# Calculate Sensitivity (True Positive Rate)
sensitivity <- tb[2, 2] / (tb[2, 2] + tb[1, 2])

# Calculate Specificity (True Negative Rate)
specificity <- tb[1, 1] / (tb[1, 1] + tb[2, 1])

# Calculate Prevalence
prevalence <- (tb[1, 2] + tb[2, 2]) / sum(tb)

# Calculate PPV (Positive Predictive Value)
ppv <- tb[2, 2] / (tb[2, 2] + tb[2, 1])

# Calculate NPV (Negative Predictive Value)
npv <- tb[1, 1] / (tb[1, 1] + tb[1, 2])

# Calculate AUROC
roc_obj <- roc(testing_set$income, probabilities)
auroc <- auc(roc_obj)

# Print the metrics
cat("accuracy:", accuracy, "\n")
cat("Sensitivity:", sensitivity, "\n")
cat("Specificity:", specificity, "\n")
cat("Prevalence:", prevalence, "\n")
cat("PPV:", ppv, "\n")
cat("NPV:", npv, "\n")
cat("AUROC:", auroc, "\n")
```

# Steven's function for statistics
# Generic Function to get all the statistics
```{r Print Model Metrics Function}

# Example:

# print_model_metrics(model = generic_model, test_data = testing_set, response_var = "income", pos_value = "Over.50K", threshold = 0.28, model_title = "Simple GLM")

#

library(caret)

library(pROC)

 

print_model_metrics <-

  function(model,

           test_data,

           response_var,

           pos_value,

           threshold = 0.5,

           model_title = "Model Metrics") {

    # Ensure that the response variable is a factor with exactly two levels

    response_factor <-

      factor(test_data[[response_var]], levels = unique(test_data[[response_var]]))

   

    # Make sure we have exactly two levels

    if (length(levels(response_factor)) != 2) {

      stop("Response variable must have exactly two levels.")

    }

   

    # Make predictions and convert to factor with the specified positive value

    probabilities <- predict(model, test_data, type = "response")

    predicted_classes <-

      factor(ifelse(

        probabilities > threshold,

        levels(response_factor)[2],

        levels(response_factor)[1]

      ),

      levels = levels(response_factor))

   

    # Calculate the confusion matrix using caret, using pos_value to specify the positive class

    cm <-

      confusionMatrix(predicted_classes, response_factor, positive = pos_value)

   

    # Extract metrics from the confusion matrix

    accuracy <- cm$overall['Accuracy']

    sensitivity <- cm$byClass['Sensitivity']

    specificity <- cm$byClass['Specificity']

    prevalence <- cm$byClass['Prevalence']

    ppv <- cm$byClass['Pos Pred Value']

    npv <- cm$byClass['Neg Pred Value']

   

    # Calculate AUROC using pROC, ensuring numeric conversion accounts for positive class being '1'

    roc_obj <-

      roc(response = as.numeric(response_factor) - 1,

          predictor = probabilities)

    auroc <- auc(roc_obj)

   

    # Replace NA values with zero or another appropriate placeholder

    metrics_values <-

      c(accuracy,

        sensitivity,

        specificity,

        prevalence,

        ppv,

        npv,

        auroc)

    metrics_values[is.na(metrics_values)] <- 0

   

    # Compile the metrics into a data frame

    metrics <- data.frame(

      Metric = c(

        "Accuracy",

        "Sensitivity",

        "Specificity",

        "Prevalence",

        "PPV",

        "NPV",

        "AUROC"

      ),

      Value = metrics_values

    )

   

    # Prepare the caption with the threshold information

    aic_val <- AIC(model)

   

    full_caption <-

      sprintf("%s\nThreshold: %s\nAIC: %.2f",

              model_title,

              threshold,

              aic_val)

   

    # Print the metrics table using knitr

    knitr::kable(

      metrics,

      col.names = c("Metric", "Value"),

      format = "markdown",

      caption = full_caption,

    )

  }

 

```


# Steven's function call
```{r}
print_model_metrics(model = full.model, test_data = testing_set_out, response_var = "income", pos_value = ">50K", threshold = 0.26, model_title = "Simple GLM")
```


# Steven's plot function
# Generic Function to plot metrix by threshold
```{r}
set.seed(1234)
library(caret)

library(pROC)

library(ggplot2)

 

# Example

# plot_model_metrics_by_threshold(model = {your_model}, test_data = {your_test_set}, response_var = "income", pos_value = {"Greater.50K"})

 

plot_model_metrics_by_threshold <-

  function(model,

           test_data,

           response_var,

           pos_value,

           thresholds = seq(0.1, 0.9, by = 0.05),

           model_title = "Model Metrics by Threshold") {

    # Ensure the response variable is a factor with exactly two levels

    response_factor <-

      factor(test_data[[response_var]], levels = unique(test_data[[response_var]]))

   

    # Make sure we have exactly two levels

    if (length(levels(response_factor)) != 2) {

      stop("Response variable must have exactly two levels.")

    }

   

    # Prepare to collect metrics

    results <-

      data.frame(

        Threshold = numeric(0),

        Accuracy = numeric(0),

        Sensitivity = numeric(0),

        Specificity = numeric(0),

        Prevalence = numeric(0),

        PPV = numeric(0),

        NPV = numeric(0)

      )

   

    # Loop over each threshold

    for (threshold in thresholds) {

      # Make predictions and convert to factor with the specified positive value

      probabilities <- predict(model, test_data, type = "response")

      predicted_classes <-

        factor(

          ifelse(

            probabilities > threshold,

            pos_value,

            levels(response_factor)[levels(response_factor) != pos_value]

          ),

          levels = levels(response_factor)

        )

     

      # Calculate the confusion matrix using caret

      cm <-

        confusionMatrix(predicted_classes, response_factor, positive = pos_value)

     

      # Store metrics

      results <- rbind(

        results,

        data.frame(

          Threshold = threshold,

          Accuracy = cm$overall['Accuracy'],

          Sensitivity = cm$byClass['Sensitivity'],

          Specificity = cm$byClass['Specificity'],

          Prevalence <- cm$byClass['Prevalence'],

          PPV = cm$byClass['Pos Pred Value'],

          NPV = cm$byClass['Neg Pred Value']

        )

      )

    }

   

    # Melt results for plotting

    results_long <-

      reshape2::melt(results, id.vars = 'Threshold', variable.name = 'Metric')

   

    # Plot results

    plot <-

      ggplot(results_long, aes(x = Threshold, y = value, color = Metric)) +

      geom_line() +

      labs(title = model_title, x = "Threshold", y = "Metric Value") +

      scale_color_manual(

        values = c(

          "Accuracy" = "blue",

          "Sensitivity" = "red",

          "Specificity" = "green",

          "Prevalence" = "purple",

          "PPV" = "orange",

          "NPV" = "brown"

        )

      ) +

      scale_y_continuous(limits = c(0, 1)) +

      ggtitle(paste("Metrics for Various Thresholds -", model_title))

   

    # Return the plot

    return(plot)

  }
```


```{r}
plot_model_metrics_by_threshold(model =full.model, test_data = testing_set_out, response_var = "income", pos_value = ">50K")
```





# Basic Stepwise: backward, forward, both
Note: There was no difference in the models suggesting the model complexity is already as simple as possibly
* The one-hot-encoding of categorical variables may also be impacting the selection process.
```{r, warning=FALSE}
cat("\n\n--------Backward Elimination---------\n\n")
# backward selection
step(full.model, trace = F, scope = list(lower=formula(full.model), upper=formula(full.model)),
     direction = 'backward')

cat("\n\n--------Forward Elimination---------\n\n")
# forward selection
step(full.model, trace = F, scope = list(lower=formula(full.model), upper=formula(full.model)),
     direction = 'forward')

cat("\n\n--------Both Elimination---------\n\n")
# forward selection
step(full.model, trace = F, scope = list(lower=formula(full.model), upper=formula(full.model)),
     direction = 'both')
```

# Stepwise for influencial outliers
```{r feature_selection, warning=FALSE}
cat("\n\n--------Backward Elimination---------\n\n")
# backward selection
step(model4, trace = F, scope = list(lower=formula(model4), upper=formula(model4)),
     direction = 'backward')

cat("\n\n--------Forward Elimination---------\n\n")
# forward selection
step(model4, trace = F, scope = list(lower=formula(model4), upper=formula(model4)),
     direction = 'forward')

cat("\n\n--------Both Elimination---------\n\n")
# forward selection
step(model4, trace = F, scope = list(lower=formula(model4), upper=formula(model4)),
     direction = 'both')
```

Reduced model that eliminates non significant variables and colinearity between those variables.
```{r}
new_formula2 <- as.formula(income ~ +age + years_education + capital_gain_log + capital_loss_log +
                             marital_status+ occupation + household_role + gender + race + hours_per_week)

new_formula2

# Fit the reduced model
reduced_model2 <- glm(new_formula2, data = training_set, family = binomial)
summary(reduced_model2)

# Check for multicollinearity
vif(reduced_model2) 

# Perform Hosmer-Lemeshow goodness of fit test
# Note: A p-value less than 0.05 typically suggests that the model does not fit the data well.
hoslem.test(training_set$income, fitted(reduced_model2))

# Make predictions
probabilities <- reduced_model2 %>% predict(testing_set, type = "response")
predicted.classes <- ifelse(probabilities > 0.27, ">50K", "<=50K")

# Model accuracy
accuracy <- mean(predicted.classes==testing_set$income)

# confusion matrix 
tb <- table(predicted.classes, testing_set$income)
tb

# Calculate Sensitivity (True Positive Rate)
sensitivity <- tb[2, 2] / (tb[2, 2] + tb[1, 2])

# Calculate Specificity (True Negative Rate)
specificity <- tb[1, 1] / (tb[1, 1] + tb[2, 1])

# Calculate Prevalence
prevalence <- (tb[1, 2] + tb[2, 2]) / sum(tb)

# Calculate PPV (Positive Predictive Value)
ppv <- tb[2, 2] / (tb[2, 2] + tb[2, 1])

# Calculate NPV (Negative Predictive Value)
npv <- tb[1, 1] / (tb[1, 1] + tb[1, 2])

# Calculate AUROC
roc_obj <- roc(testing_set$income, probabilities)
auroc <- auc(roc_obj)

# Print the metrics
cat("accuracy:", accuracy, "\n")
cat("Sensitivity:", sensitivity, "\n")
cat("Specificity:", specificity, "\n")
cat("Prevalence:", prevalence, "\n")
cat("PPV:", ppv, "\n")
cat("NPV:", npv, "\n")
cat("AUROC:", auroc, "\n")



############### Outliers Removed #############################

# Fit the reduced model
reduced_model3 <- glm(new_formula2, data = training_set_out, family = binomial)
summary(reduced_model2)

# Check for multicollinearity
vif(reduced_model3) 

# Perform Hosmer-Lemeshow goodness of fit test
# Note: A p-value less than 0.05 typically suggests that the model does not fit the data well.
hoslem.test(training_set_out$income, fitted(reduced_model3))

# Make predictions
probabilities2 <- reduced_model3 %>% predict(testing_set_out, type = "response")
predicted.classes2 <- ifelse(probabilities2 > 0.27, ">50K", "<=50K")

# Model accuracy
accuracy <- mean(predicted.classes2==testing_set_out$income)

# confusion matrix 
tb2 <- table(predicted.classes2, testing_set_out$income)
tb2

# Calculate Sensitivity (True Positive Rate)
sensitivity2 <- tb2[2, 2] / (tb2[2, 2] + tb2[1, 2])

# Calculate Specificity (True Negative Rate)
specificity2 <- tb2[1, 1] / (tb2[1, 1] + tb2[2, 1])

# Calculate Prevalence
prevalence2 <- (tb2[1, 2] + tb2[2, 2]) / sum(tb2)

# Calculate PPV (Positive Predictive Value)
ppv2 <- tb2[2, 2] / (tb2[2, 2] + tb2[2, 1])

# Calculate NPV (Negative Predictive Value)
npv2 <- tb2[1, 1] / (tb2[1, 1] + tb2[1, 2])

# Calculate AUROC
roc_obj2 <- roc(testing_set_out$income, probabilities2)
auroc2 <- auc(roc_obj)

# Print the metrics
cat("accuracy:", accuracy2, "\n")
cat("Sensitivity:", sensitivity2, "\n")
cat("Specificity:", specificity2, "\n")
cat("Prevalence:", prevalence2, "\n")
cat("PPV:", ppv2, "\n")
cat("NPV:", npv2, "\n")
cat("AUROC:", auroc2, "\n")


# full with outliers
pred_full <- prediction(probabilities, testing_set$income)
pred_full_per <- performance(pred_full, measure = "tpr", x.measure = "fpr")
mod1_pred <- data.frame(FP = pred_full_per@x.values[[1]], TP = pred_full_per@y.values[[1]])

# full without outliers
pred_full_Out <- prediction(probabilities2, testing_set_out$income)
pred_full_per2 <- performance(pred_full_Out, measure = "tpr", x.measure = "fpr")
mod2_pred <- data.frame(FP = pred_full_per2@x.values[[1]], TP = pred_full_per2@y.values[[1]])

# plot ROC curve for logistic regression
g <- ggplot() + 
  geom_line(data = mod1_pred, aes(x = FP, y = TP, color = 'Reduced With Outliers')) + 
  geom_line(data = mod2_pred, aes(x = FP, y = TP, color = 'Reduced Without Outliers')) +
  geom_segment(aes(x = 0, xend = 1, y = 0, yend = 1)) +
  ggtitle('ROC Curve') + 
  labs(x = 'False Positive Rate', y = 'True Positive Rate') 


g +  scale_colour_manual(name = 'Classifier', values = c('Reduced With Outliers'='red', 
                                               'Reduced Without Outliers'='blue'))
```


# Odds ratio for the reduced model removing influencial outliers.
An OR of x suggests that with each additional year of (the variable), the odds of the outcome occurring increase by approximately ((OR -1)*100) all other variables remaining constant. or if it has a reference variable...compared to the reference variable.
```{r warning=FALSE}
summary(reduced_model3)
# Calculate Odds Ratios and 90% Confidence Intervals
odds_ratios <- exp(coef(reduced_model3))
conf_int <- exp(confint(reduced_model3, level = 0.95))

# Identify significant predictors at the 0.05 level
significant_predictors <- summary(reduced_model3)$coefficients[, 4] < 0.05

# Extract Odds Ratios and Confidence Intervals for significant predictors
# Note: Use names() to match significant predictors by name
significant_or_ci <- data.frame(
  OddsRatio = odds_ratios[significant_predictors],
  LowerCI = conf_int[names(odds_ratios)[significant_predictors], 1],
  UpperCI = conf_int[names(odds_ratios)[significant_predictors], 2]
)
significant_or_ci

```



# Most of the residuals fall within +/- 3
```{r}
# create a data frame to store information regarding deviance residuals
index <- 1:dim(training_set)[1]
dev_resid <- residuals(model3)
income <- training_set$income
dff <- data.frame(index, dev_resid, income)

ggplot(dff, aes(x = index, y = dev_resid, color = income)) +
  geom_point() + 
  geom_hline(yintercept = 3, linetype = 'dashed', color = 'blue') +
  geom_hline(yintercept = -3, linetype = 'dashed', color = 'blue')
```
# Objective two Models
# Test with interactions added on suspected variables with higher multicollinearity using the One-Hot_encoded data. Some interaction showed significance, but the model performed equal to model 3 above.
```{r}
################## glm test with influential outliers removed ##############################
cat("\n\n----------------interactions--------------\n\n")

interactions_formula <- as.formula(income ~ +age + workclass_num + education_level_num + years_education + marital_status_num+ 
                              occupation_num + household_role_num + race_num + gender_num + capital_gain_log + capital_loss_log +  hours_per_week +
                                native_country_num + capital_gain_log:capital_loss_log + 
                               education_level_num:years_education +workclass_num:occupation_num)
interactions_formula

interactions_formula2 <- as.formula(income ~ +age + workclass_num + similar_pop_count + education_level_num + years_education + marital_status_num+ 
                              occupation_num + household_role_num + race_num + gender_num + capital_gain + capital_loss +  hours_per_week +
                                native_country_num + invested_num + capital_gain:capital_loss + 
                               education_level_num:years_education +workclass_num:occupation_num)

# Fit a logistic regression model
model6 <- glm(interactions_formula, data = training_set_LR, family = binomial)

# Confidence interval
# confint(model3)

# Summarize the final selected model
summary(model6)

# Check for multicollinearity
# vif(model6) 

# Perform Hosmer-Lemeshow goodness of fit test
# Note: A p-value less than 0.05 typically suggests that the model does not fit the data well.
hoslem.test(training_set_LR$income, fitted(model6))

# Make predictions
probabilities <- model6 %>% predict(testing_set_LR, type = "response")
predicted.classes <- ifelse(probabilities > 0.28, "1", "0")

# Model accuracy
accuracy <- mean(predicted.classes==testing_set_LR$income)

# confusion matrix 
tb <- table(predicted.classes, testing_set_LR$income)
tb

# Calculate Sensitivity (True Positive Rate)
sensitivity <- tb[2, 2] / (tb[2, 2] + tb[1, 2])

# Calculate Specificity (True Negative Rate)
specificity <- tb[1, 1] / (tb[1, 1] + tb[2, 1])

# Calculate Prevalence
prevalence <- (tb[1, 2] + tb[2, 2]) / sum(tb)

# Calculate PPV (Positive Predictive Value)
ppv <- tb[2, 2] / (tb[2, 2] + tb[2, 1])

# Calculate NPV (Negative Predictive Value)
npv <- tb[1, 1] / (tb[1, 1] + tb[1, 2])

# Calculate AUROC
roc_obj <- roc(testing_set_LR$income, probabilities)
auroc <- auc(roc_obj)

# Print the metrics
cat("accuracy:", accuracy, "\n")
cat("Sensitivity:", sensitivity, "\n")
cat("Specificity:", specificity, "\n")
cat("Prevalence:", prevalence, "\n")
cat("PPV:", ppv, "\n")
cat("NPV:", npv, "\n")
cat("AUROC:", auroc, "\n")
```


#Joel
```{r lda}
library(caret)
# change the income value names to work with glmnet (greater than sign will not run)
training_set2 <- training_set
training_set2$income <- gsub('<=50K', 'less_than_50K', training_set2$income)
training_set2$income <- gsub('>50K', 'greater_than_50K', training_set2$income)

testing_set2 <- testing_set
testing_set2$income <- gsub('<=50K', 'less_than_50K', testing_set2$income)
testing_set2$income <- gsub('>50K', 'greater_than_50K', testing_set2$income)

levels(training_set2$income)

fitControl<-trainControl(method="repeatedcv",number=2,repeats=1,classProbs=TRUE, summaryFunction=mnLogLoss)
set.seed(1234)

# Convert 'income' to a factor
training_set2$income <- factor(training_set2$income, levels = c('greater_than_50K', 'less_than_50K'))
testing_set2$income <- factor(testing_set2$income, levels = c('greater_than_50K', 'less_than_50K'))



# Fit a logistic regression model
logreg_model <- train(income ~ ., data = training_set2,
                      method = "lda",
                      trControl = fitControl,
                      metric = "logLoss")

# Compute predicted probabilities on the training data
logreg_predictions <- predict(logreg_model, testing_set2, type = "prob")[, 'less_than_50K']


threshold <- 0.78
# Create factor predictions based on a threshold
logreg_preds <- factor(ifelse(logreg_predictions > threshold, 'less_than_50K', 'greater_than_50K' ),
                       levels = c( 'less_than_50K', 'greater_than_50K' ))

# Get the confusion matrix
confusionMatrix(data = logreg_preds, reference = testing_set2$income)

library(ROCR)

roc_obj <- roc(testing_set2$income, logreg_predictions)
auroc <- auc(roc_obj)

cat("AUROC:", auroc, "\n")




```


# Non-Parametric Model testing.

# feedforward neural network (NN) with one hidden layer
* Sources: https://rpubs.com/H_Zhu/235617
* https://rpubs.com/mbaumer/NeuralNetworks
```{r nn1}
set.seed(1234)
library(nnet)
library(NeuralNetTools)

# run the model
nn1 <- nnet(income ~ ., data = training_set_out, size = 20, maxit = 500)
# summary(nn1)

# predict against the test set
nn1.pred <- predict(nn1, newdata = testing_set_out, type = 'raw')

pred1 <- rep('<=50K', length(nn1.pred))
pred1[nn1.pred>=.28] <- '>50K' # Threshold set here

# Create a neural network graph
plotnet(nn1, y_names = "IncomeLevel")
title("Graphical Representation of our Neural Network")

# Feature selection
garson(nn1)
garson(nn1, bar_plot = FALSE)

# confusion matrix 
tb1 <- table(pred1, testing_set_out$income)
tb1

# Model accuracy - overall correctness
accuracy <- mean(pred1==testing_set_out$income)

# Calculate Sensitivity (True Positive Rate)
sensitivity <- tb1[2, 2] / (tb1[2, 2] + tb1[1, 2])

# Calculate Specificity (True Negative Rate)
specificity <- tb1[1, 1] / (tb1[1, 1] + tb1[2, 1])

# Calculate Prevalence - percentage of the positive term in the dataset
prevalence <- (tb1[1, 2] + tb1[2, 2]) / sum(tb1)

# Calculate PPV (Positive Predictive Value)
ppv <- tb1[2, 2] / (tb1[2, 2] + tb1[2, 1])

# Calculate NPV (Negative Predictive Value)
npv <- tb1[1, 1] / (tb1[1, 1] + tb1[1, 2])

# Calculate AUROC
roc_obj <- roc(testing_set_out$income, nn1.pred)
auroc <- auc(roc_obj)

# Print the metrics
cat("accuracy:", accuracy, "\n")
cat("Sensitivity:", sensitivity, "\n")
cat("Specificity:", specificity, "\n")
cat("Prevalence:", prevalence, "\n")
cat("PPV:", ppv, "\n")
cat("NPV:", npv, "\n")
cat("AUROC:", auroc, "\n")
```

Classification and Regression Trees (CART) model is obtained by recursively partitioning the data space and fitting a simple prediction model within each partition. 
* It evaluates all possible splits and selects the one that best reduces impurity (e.g., Gini impurity for classification or residual reduction for regression).
* To prevent overfitting, pruning techniques are applied to remove nodes that contribute little to model accuracy.

```{r}
library(rpart)
library(rpart.plot)

new_formula2 <- as.formula(income ~ +age + years_education + capital_gain_log + capital_loss_log +
                             marital_status+ occupation + household_role + gender + hours_per_week)

set.seed(1234)
# Build the decision tree model # cp is the complexity tradeoff - smaller is better
tree2 <- rpart(new_formula2, data = training_set_out, method = 'class', cp = 1e-20) 

# Plot the decision tree 
rpart.plot(tree2, fallen.leaves = FALSE, tweak = 1.1, varlen = 4, faclen = 4)
rpart.plot(tree2)

# Predict class probabilities
tree2.pred.prob <- predict(tree2, newdata = testing_set_out, type = 'prob')

# Apply threshold to get predicted classes
threshold <- 0.22
tree2.pred <- ifelse(tree2.pred.prob[, ">50K"] > threshold, ">50K", "<=50K")

# Confusion matrix
tb2 <- table(tree2.pred, testing_set_out$income)
tb2
# Model accuracy
accuracy <- mean(tree2.pred == testing_set_out$income)

# Calculate Sensitivity (True Positive Rate)
sensitivity <- tb2[2, 2] / (tb2[2, 2] + tb2[1, 2])

# Calculate Specificity (True Negative Rate)
specificity <- tb2[1, 1] / (tb2[1, 1] + tb2[2, 1])

# Calculate Prevalence
prevalence <- (tb2[1, 2] + tb2[2, 2]) / sum(tb2)

# Calculate PPV (Positive Predictive Value)
ppv <- tb2[2, 2] / (tb2[2, 2] + tb2[2, 1])

# Calculate NPV (Negative Predictive Value)
npv <- tb2[1, 1] / (tb2[1, 1] + tb2[1, 2])

# Assuming the positive class is '>50K'
positive_class_probs <- tree2.pred.prob[, ">50K"]

# Calculate AUROC using the vector of probabilities for the positive class
roc_obj <- roc(testing_set_out$income, positive_class_probs)
auroc <- auc(roc_obj)

# Print the metrics
cat("Accuracy:", accuracy, "\n")
cat("Sensitivity:", sensitivity, "\n")
cat("Specificity:", specificity, "\n")
cat("Prevalence:", prevalence, "\n")
cat("PPV:", ppv, "\n")
cat("NPV:", npv, "\n")
cat("AUROC:", auroc, "\n")

```


# Classification and Regression Trees (CART)
# Source: https://carpentries-incubator.github.io/r-ml-tabular-data/04-Decision-Forests/index.html
```{r decision_tree}
set.seed(1234)
library(rpart)
# install.packages("rpart.plot")
library(rpart.plot)

# create a decision tree model
rwtree <- rpart(new_formula2, data = training_set_out, method = "class", cp = 1e-10) # CP Complexity- lower is more complex
rpart.plot(rwtree)
# summary(rwtree)

# Predict class probabilities
rwtree.pred.prob <- predict(rwtree, newdata = testing_set_out, type = 'prob')

# Apply threshold to get predicted classes
threshold <- 0.22
rwtree.pred <- ifelse(rwtree.pred.prob[, ">50K"] > threshold, ">50K", "<=50K")

# Confusion matrix
tb5 <- table(rwtree.pred, testing_set_out$income)
tb5

rwtreeErrors <- rwtree.pred.prob - ifelse(testing_set_out$income == ">50K", 1, 0)
rwtreeRMSE <- sqrt(mean(rwtreeErrors^2))



# Model accuracy
accuracy <- mean(rwtree.pred == testing_set_out$income)

# Calculate Sensitivity (True Positive Rate)
sensitivity <- tb5[2, 2] / (tb5[2, 2] + tb5[1, 2])

# Calculate Specificity (True Negative Rate)
specificity <- tb5[1, 1] / (tb5[1, 1] + tb5[2, 1])

# Calculate Prevalence
prevalence <- (tb5[1, 2] + tb5[2, 2]) / sum(tb5)

# Calculate PPV (Positive Predictive Value)
ppv <- tb5[2, 2] / (tb5[2, 2] + tb5[2, 1])

# Calculate NPV (Negative Predictive Value)
npv <- tb5[1, 1] / (tb5[1, 1] + tb5[1, 2])

# Assuming the positive class is '>50K'
positive_class_probs <- rwtree.pred.prob[, ">50K"]

# Calculate AUROC using the vector of probabilities for the positive class
roc_obj <- roc(testing_set_out$income, positive_class_probs)
auroc <- auc(roc_obj)

# Print the metrics
cat("Accuracy:", accuracy, "\n")
cat("Sensitivity:", sensitivity, "\n")
cat("Specificity:", specificity, "\n")
cat("Prevalence:", prevalence, "\n")
cat("PPV:", ppv, "\n")
cat("NPV:", npv, "\n")
cat("AUROC:", auroc, "\n")
cat("RMSE:", rwtreeRMSE, "\n")
```


# Random forest (RF) improves predictive accuracy by generating a large number of bootstrapped trees.
# Sources: https://carpentries-incubator.github.io/r-ml-tabular-data/04-Decision-Forests/index.html
* MeanDecreaseAccuracy measures how much the model’s accuracy decreases when a specific feature is excluded. Higher values indicate that excluding that feature would significantly impact prediction accuracy.
* MeanDecreaseGini assesses how each feature contributes to the homogeneity of nodes and leaves in the random forest.A higher value suggests that excluding that feature would lead to less pure (more impure) nodes.
* Compare the importance rankings between the two metrics. If a feature is consistently high in both panels, it is likely crucial.
* Plot - Out-of-Bag (OOB) Error Line (Green, Dotted), Test Error Line (Red, Dashed): both models plateau quickly and are no longer viable after 200 trees.
```{r}
set.seed(1234)
library(randomForest)

# Train the random forest model
rf3t <- randomForest(income ~ ., data = training_set_out, ntree = 300, keep.forest=TRUE, importance=TRUE,test=testing_set_out$income)
rf3t
plot(rf3t)

# Visually check variable importance.
importance(rf3t)
varImpPlot(rf3t)


# Predict class probabilities
rf3t.pred.prob <- predict(rf3t, newdata = testing_set_out, type = 'prob')

# Apply threshold to get predicted classes
threshold <- 0.2
rf3t.pred <- ifelse(rf3t.pred.prob[, ">50K"] > threshold, ">50K", "<=50K")

# Confusion matrix
tb3t <- table(rf3t.pred, testing_set_out$income)
tb3t

# Get the RMSE -not sure how useful this is because it is categorical
rfErrors <- rf3t.pred.prob - ifelse(testing_set_out$income == ">50K", 1, 0)
rfRMSE <- sqrt(mean(rfErrors^2))
rfRMSE

# Model accuracy
accuracy <- mean(rf3t.pred == testing_set_out$income)

# Calculate Sensitivity (True Positive Rate)
sensitivity <- tb3t[2, 2] / (tb3t[2, 2] + tb3t[1, 2])

# Calculate Specificity (True Negative Rate)
specificity <- tb3t[1, 1] / (tb3t[1, 1] + tb3t[2, 1])

# Calculate Prevalence
prevalence <- (tb3t[1, 2] + tb3t[2, 2]) / sum(tb3t)

# Calculate PPV (Positive Predictive Value)
ppv <- tb3t[2, 2] / (tb3t[2, 2] + tb3t[2, 1])

# Calculate NPV (Negative Predictive Value)
npv <- tb3t[1, 1] / (tb3t[1, 1] + tb3t[1, 2])

# Assuming the positive class is '>50K'
positive_class_probs <- rf3t.pred.prob[, ">50K"]

# Calculate AUROC using the vector of probabilities for the positive class
roc_obj <- roc(testing_set_out$income, positive_class_probs)
auroc <- auc(roc_obj)

# Print the metrics
cat("Accuracy:", accuracy, "\n")
cat("Sensitivity:", sensitivity, "\n")
cat("Specificity:", specificity, "\n")
cat("Prevalence:", prevalence, "\n")
cat("PPV:", ppv, "\n")
cat("NPV:", npv, "\n")
cat("AUROC:", auroc, "\n")
cat("RMSE:", rfRMSE, "\n")

```

# Random Forest reduced model
```{r RF_reduced}
set.seed(1234)
cat("\n---------------reduced model-------------\n")

new_formula2 <- as.formula(income ~ +age + years_education + capital_gain_log + capital_loss_log +
                             marital_status+ occupation + household_role + hours_per_week)

# Train the reduced random forest model
rf3 <- randomForest(new_formula2, data = training_set_out, ntree = 300)
rf3
plot(rf3)

# Visually check variable importance.
importance(rf3)
varImpPlot(rf3)


# Predict class probabilities
rf3.pred.prob <- predict(rf3, newdata = testing_set_out, type = 'prob')

# Apply threshold to get predicted classes
threshold <- 0.13
rf3.pred <- ifelse(rf3.pred.prob[, ">50K"] > threshold, ">50K", "<=50K")

# Confusion matrix
tb3 <- table(rf3.pred, testing_set_out$income)
tb3

# Get the RMSE -not sure how useful this is because it is categorical
rfErrors <- rf3.pred.prob - ifelse(testing_set_out$income == ">50K", 1, 0)
rfRMSE <- sqrt(mean(rfErrors^2))

# Model accuracy
accuracy <- mean(rf3.pred == testing_set_out$income)

# Calculate Sensitivity (True Positive Rate)
sensitivity <- tb3[2, 2] / (tb3[2, 2] + tb3[1, 2])

# Calculate Specificity (True Negative Rate)
specificity <- tb3[1, 1] / (tb3[1, 1] + tb3[2, 1])

# Calculate Prevalence
prevalence <- (tb3[1, 2] + tb3[2, 2]) / sum(tb3)

# Calculate PPV (Positive Predictive Value)
ppv <- tb3[2, 2] / (tb3[2, 2] + tb3[2, 1])

# Calculate NPV (Negative Predictive Value)
npv <- tb3[1, 1] / (tb3[1, 1] + tb3[1, 2])

# Assuming the positive class is '>50K'
positive_class_probs <- rf3.pred.prob[, ">50K"]

# Calculate AUROC using the vector of probabilities for the positive class
roc_obj <- roc(testing_set_out$income, positive_class_probs)
auroc <- auc(roc_obj)

# Print the metrics
cat("Accuracy:", accuracy, "\n")
cat("Sensitivity:", sensitivity, "\n")
cat("Specificity:", specificity, "\n")
cat("Prevalence:", prevalence, "\n")
cat("PPV:", ppv, "\n")
cat("NPV:", npv, "\n")
cat("AUROC:", auroc, "\n")
cat("RMSE:", rfRMSE, "\n")

```


# Support vector machine (SVM) to predict the income level
```{r}
library(kernlab)
library(caret)

set.seed(1234)
# Train the SVM model
svm4 <- ksvm(income ~ ., data = training_set_out, )

# Get decision values
svm4.pred.prob <- predict(svm4, newdata = testing_set_out, type = 'decision')

# Apply threshold to decision values to get predicted classes
threshold <- -.9 # Adjust this threshold (not normal for SVM)
svm4.pred <- ifelse(svm4.pred.prob > threshold, ">50K", "<=50K")


# Confusion matrix
tb4 <- table(svm4.pred, testing_set_out$income)
tb4
# Calculate accuracy and other metrics
accuracy <- mean(svm4.pred == testing_set_out$income)
sensitivity <- tb4[2, 2] / (tb4[2, 2] + tb4[1, 2])
specificity <- tb4[1, 1] / (tb4[1, 1] + tb4[2, 1])
prevalence <- (tb4[1, 2] + tb4[2, 2]) / sum(tb4)
ppv <- tb4[2, 2] / (tb4[2, 2] + tb4[2, 1])
npv <- tb4[1, 1] / (tb4[1, 1] + tb4[1, 2])

# Calculate AUROC using the decision values
roc_obj <- roc(testing_set_out$income, svm4.pred.prob)
auroc <- auc(roc_obj)

# Print the metrics
cat("Accuracy:", accuracy, "\n")
cat("Sensitivity:", sensitivity, "\n")
cat("Specificity:", specificity, "\n")
cat("Prevalence:", prevalence, "\n")
cat("PPV:", ppv, "\n")
cat("NPV:", npv, "\n")
cat("AUROC:", auroc, "\n")

```




* Accuracy: This tells you the overall correctness of the model, but it doesn’t account for the class imbalance or the costs of different types of errors.
* Sensitivity (True Positive Rate): Indicates how well the model predicts the positive class, in this case, those earning over 50K. It’s important if you don’t want to miss individuals who are actually earning more than 50K.
*Specificity (True Negative Rate): Shows how well the model predicts the negative class, those earning under 50K. It’s important if you want to minimize false alarms, where people are incorrectly predicted to earn more than 50K.
* Prevalence: This is the actual proportion of the positive class in your dataset. It doesn’t help with prediction but gives context about the dataset.
* PPV (Positive Predictive Value): Tells you how likely it is that a person predicted to earn over 50K actually does. This is useful if the cost of falsely identifying someone as earning over 50K is high.
* NPV (Negative Predictive Value): Indicates the likelihood that a person predicted to earn under 50K actually earns less than that. It’s useful when it’s important not to misclassify those who earn under 50K.
* AUROC (Area Under the Receiver Operating Characteristic curve): This metric is useful for evaluating the model’s ability to discriminate between the two classes across all thresholds. A high AUROC indicates a model that is good at ranking predictions rather than making a hard cutoff.

# ROC Curve
Sources: https://www.statology.org/interpret-roc-curve/

```{r}
prob <- predict(model4, testing_set, type = 'response') # reduced
# create a prediction object
pr <- prediction(prob, testing_set$income)
prf <- performance(pr, measure = "tpr", x.measure = "fpr")

# create a data frame for TP and FP rates
dd <- data.frame(FP = prf@x.values[[1]], TP = prf@y.values[[1]])

# complex LR
prob5 <- predict(model6, testing_set_LR, type = 'response') #interactions
# create a prediction object
pr5 <- prediction(prob5, testing_set_LR$income)
prf5 <- performance(pr5, measure = "tpr", x.measure = "fpr")

# create a data frame for TP and FP rates
dd5 <- data.frame(FP = prf5@x.values[[1]], TP = prf5@y.values[[1]])

# NN
pr1 <- prediction(nn1.pred, testing_set_out$income)
prf1 <- performance(pr1, measure = "tpr", x.measure = "fpr")
dd1 <- data.frame(FP = prf1@x.values[[1]], TP = prf1@y.values[[1]])

# CART
pr2 <- prediction(tree2.pred.prob[,2], testing_set_out$income)
prf2 <- performance(pr2, measure = "tpr", x.measure = "fpr")
dd2 <- data.frame(FP = prf2@x.values[[1]], TP = prf2@y.values[[1]])

# RF
pr3 <- prediction(rf3.pred.prob[,2], testing_set_out$income)
prf3 <- performance(pr3, measure = "tpr", x.measure = "fpr")
dd3 <- data.frame(FP = prf3@x.values[[1]], TP = prf3@y.values[[1]])

# SVM
pr4 <- prediction(svm4.pred.prob, testing_set_out$income)
prf4 <- performance(pr4, measure = "tpr", x.measure = "fpr")
dd4 <- data.frame(FP = prf4@x.values[[1]], TP = prf4@y.values[[1]])

# Create a prediction object for the lda model (logreg_predictions)
pr6 <- prediction(logreg_predictions, testing_set2$income)

# Calculate the True Positive Rate (TPR) and False Positive Rate (FPR)
prf6 <- performance(pr6, measure = "tpr", x.measure = "fpr")
dd6 <- data.frame(FP = prf6@x.values[[1]], TP = prf6@y.values[[1]])



# plot ROC curve for logistic regression
g <- ggplot() + 
  geom_line(data = dd, aes(x = FP, y = TP, color = 'Logistic Regression')) + 
  geom_line(data = dd5, aes(x = FP, y = TP, color = 'Complex Logistic Regression')) + 
  geom_line(data = dd1, aes(x = FP, y = TP, color = 'Neural Networks')) + 
  geom_line(data = dd2, aes(x = FP, y = TP, color = 'CART')) + 
  geom_line(data = dd3, aes(x = FP, y = TP, color = 'Random Forest')) +
  geom_line(data = dd4, aes(x = FP, y = TP, color = 'Support Vector Machine')) +
  geom_line(data = dd6, aes(x = FP, y = TP, color = 'Linear Discriminant Analysis')) +
  geom_segment(aes(x = 0, xend = 1, y = 0, yend = 1)) +
  ggtitle('ROC Curve') + 
  labs(x = 'False Positive Rate', y = 'True Positive Rate') 


g +  scale_colour_manual(name = 'Classifier', values = c('Logistic Regression'='#E69F00', 
                                                         'Complex Logistic Regression'='red',
                                               'Neural Networks'='blue', 'CART'='#009E73', 
                                               'Random Forest'='#D55E00', 'Support Vector Machine'='#0072B2',
                                               'Linear Discriminant Analysis' = 'green'))
```


```{r}
# AUC
auc <- rbind(performance(pr, measure = 'auc')@y.values[[1]],
             performance(pr1, measure = 'auc')@y.values[[1]],
             performance(pr2, measure = 'auc')@y.values[[1]],
             performance(pr3, measure = 'auc')@y.values[[1]],
             performance(pr4, measure = 'auc')@y.values[[1]],
             performance(pr5, measure = 'auc')@y.values[[1]],
             performance(pr6, measure = 'auc')@y.values[[1]])
rownames(auc) <- (c('Logistic Regression', 'Neural Networks', 'CART',
                                    'Random Forest', 'Support Vector Machine', 
                    'Complex Logistic Regression', 'Linear Discriminant Analysis'))
colnames(auc) <- 'Area Under ROC Curve'
round(auc, 4)
```

# Appendix








